---
title: "Exploratory Data Analysis 2"
subtitle: "R Notes"
author: "Luc Anselin and Grant Morrison^[University of Chicago, Center for Spatial Data Science -- anselin@uchicago.edu,morrisonge@uchicago.edu]"
date: "latest update 09/17/2018"
output:
  html_document:
    fig_caption: yes
    self_contained: no
    toc: yes
    toc_depth: 4
    css: tutor.css
references:
- id: scatterplot3d03
  title: Scatterplot3d - an R package for Visualizing Multivariate Data
  author:
  - family: Ligges
    given: Uwe
  - family: Machler
    given: Martin
  container-title: Journal of Statistical Software
  volume: 8
  issued:
    year: 2003
  type: article-journal
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

## Introduction {-}

This notebook cover the functionality of the [Exploratory Data Analysis 2](https://geodacenter.github.io/workbook/2b_eda_multi/lab2b.html) section of the GeoDa workbook. We refer to that document for details on the methodology, references, etc. The goal of these notes is to approximate as closely as possible the operations carried out using GeoDa by means of a range of R packages.

The notes are written with R beginners in mind, more seasoned R users can probably skip most of the comments
on data structures and other R particulars. Also, as always in R, there are typically several ways to achieve a specific objective, so what is shown here is just one way that works, but there often are others (that may even be more elegant, work faster, or scale better).

For this notebook, we continue to use the socioeconomic data about 55 sub-boroughs in NYC from the GeoDa website. Our goal in this lab is show how to implement exploratory data analysis methods with three or more variables. 

### Objectives {-}

After completing the notebook, you should know how to carry out the following tasks:

- Creating a scatterplot matrix

- Adding different types of smoothers to a scatter plot matrix

- Creating a bubble plot

- Creating a 3d scatter plot

- Creating a parallel coordinate plot

- Constructing conditional plots 

- Making graphs interactive

#### R Packages used {-}

- **tidyverse**: for general data wrangling (includes **readr** and **dplyr**)

- **ggplot2**: to draw statistical plots, including conditional plots. We use this rather than base R for increased functionality and more aesthetically pleasing plots (included in **tidyverse**)

- **GGally**: a **ggplot** add-on package to create a scatterplot matrix and parallel coordinate plot

- **scatterplot3d**: to create a static 3d scattter plot

- **plotly**: to construct interactive 3d scatter and parallel coordinate plots

#### R Commands used {-}

Below follows a list of the commands used in this notebook. For further details
and a comprehensive list of options, please consult the 
[R documentation](https://www.rdocumentation.org).

- **Base R**: `setwd`, `install.packages`, `library`, `head`, `names`

- **tidyverse**: `read_csv`, `rename`, `select`

- **GGally**: `ggscatmat`, `ggpairs`, `ggparcoord`

- **ggplot2**: `ggplot`, `geom_point`, `xlab`, `ylab`, `ggtitle`, `theme`,
     `cut_number`, `facet_grid`, `geom_smooth`, `geom_histogram`

- **scatterplot3d**: `scatterplot3d`

- **plotly**: `plot_ly`, `add_markers`, `layout`

## Preliminaries {-}

Before starting, make sure to have the latest version of R and of packages that are compiled for the matching version of R (this document was created using R 3.5.1 of 2018-07-02). Also, make sure to set a working directory, such that the
data set is in the right path.^[Use `setwd(directorypath)` to specify the working directory.]

### Load packages {-}

First, we load all the required packages using the `library` command. If you don't have some of these in your system, make sure to install them first as well as
their dependencies.^[Use 
`install.packages(packagename)`.] You will get an error message if something is missing. If needed, just install the missing piece and everything will work after that. Note
that **ggplot2** does not need to be loaded separately since it is included
in the **tidyverse** package collection.

```{r}
library(tidyverse)
library(GGally)
library(scatterplot3d)
library(plotly)

```

### Obtaining the data {-}

The data to implement the operations in this workbook are contained in [NYC Data](https://geodacenter.github.io/data-and-lab/nyc/) on the GeoDa support
web site. After the file is downloaded, it must be unzipped (e.g., double click 
on the file). The **nyc** folder should be moved to the current working directory
for the path names we use below to work correctly.

#### Creating an initial data frame {-}
We use the **tidyverse** function `read_csv` to read the data into a data frame
**nyc.data**. We could also have used the base R `read.csv`, but `read_csv` is a bit more robust and creates a **tibble**, a data frame with some additional information.
As usual, we check the contents of the data frame with a `head` command.

```{r}
nyc.data <- read_csv("nyc/nyc.csv")
head(nyc.data)
```

#### Making the variable names compatible {-}
As in the previous exercise, we need to make the variable names compatible
with their lower case counterparts in the GeoDa Workbook. Again, we will use 
the **tidyverse** `rename` function to turn the all-caps variables into
lower case for the examples we will use. As in the GeoDa workbook, we will use
the average people per household in 2000 (**hhsiz00**), the percentage households with children under 18 in 2000 (**kids2000**), the average number of years lived in the current residence in 2002 (**yrhom02**), the percentage households receiving public assistance in 2000 (**pubast00**), and the median rent in 2002 (**rent2002**).

```{r}
nyc.data <- nyc.data %>% rename("hhsiz00" = "HHSIZ00","kids2000" = "KIDS2000",
                                "yrhom02"="YRHOM02","pubast00" = "PUBAST00",
                                "rent2002"="RENT2002")
names(nyc.data)
```


## Scatter Plot Matrix {-}

A scatter plot matrix visualizes the bivariate relationships among several pairs of variables. The individual scatter plots are stacked such that each variable is in turn on the x-axis and on the y-axis. 

### Basic scatter plot matrix {-}
A scatter plot matrix is not included in the functionality of **ggplot2**, but it can
be created in a number of ways using the **GGally** package, which 
extends **ggplot2** with many additional features. Extensive documentation of **GGally** functionality is available
on its [Github page](http://ggobi.github.io/ggally/).

A quick and dirty scatter plot matrix is created by means of the `ggscatmat` command
(detailed documentation is available on the [GGally Github page](http://ggobi.github.io/ggally/rd.html#ggscatmat)).

The `ggscatmat` function provides pairwise scatter plots in a lower diagonal of the graph, a density graph in the diagonal, and the pairwise correlations in the upper diagonal. This contrasts
with GeoDa, where all pairwise scatter plots are given, and the diagonal is populated
with a histogram for the individual variable (GeoDa currently does not support density plots).

The command is very simple: it takes the data set, the list of variables passed to
`columns` and a few options (color, choice of correlation coefficient). However, it does not seem to contain a way to show a linear or nonlinear smoother. In order
to accomplish this, we will need the more powerful `ggpairs` function (see below).

We follow the example in the GeoDa Workbook and use the four variables 
**hhsiz00**, **kids2000**, **yrhom02**, and **pubast00**.

```{r}
ggscatmat(nyc.data, columns= c("hhsiz00","kids2000", "yrhom02", "pubast00"))
```

### Scatter plot matrix with smoothing {-}
An alternative approach that provides much finer control of the graph can
be based on the `ggpairs` function of **GGally** (see 
[the GGobi Github page](http://ggobi.github.io/ggally/rd.html#ggpairs) for extensive documentation). As in `ggscatmat`, this
function takes the data set as an argument, followed by the variables
specified in the `columns` argument. The lower and upper triangle part of the
matrix, and the diagonal are specified by means of the arguments `lower`, `upper`,
and `diag`. The values for these arguments must be passed as a `list`.

#### Default scatter plot matrix {-}

First, we illustrate the default setting. We do not need to include the
specifics, but they amount to:

- `lower=list(continuous="points")`, for a scatter plot in the lower triangle

- `diag=list(continuous="densityDiag")`, for a density plot on the diagonal

- `upper=list(coninuous='cor')`, for a correlation coefficient in the upper diagonal

```{r}
ggpairs(nyc.data, columns=c("hhsiz00","kids2000", "yrhom02", "pubast00"))
```

#### Pairwise scatter plots {-}
In order to obtain a scatter plot in both lower and upper triangles, we set
`upper = list(continuous="points")` in the arguments to **ggpairs**. In addition,
to have histograms in the diagonal, set set `diag=list(continuous="barDiag")`.

```{r}
ggpairs(nyc.data, columns=c("hhsiz00","kids2000", "yrhom02", "pubast00"),
        upper=list(continuous="points"),diag=list(continuous="barDiag"))
```

#### Scatter plot matrix with linear smoother {-}
At this point, we can add a linear smoother by specifying 
`list(continuous="smooth")` instead of `continuous="points"` for both
`upper` and `lower` parameters (back with the default density plot on the
diagonal).

```{r}
ggpairs(nyc.data, columns=c("hhsiz00","kids2000", "yrhom02", "pubast00"),
        upper=list(continuous="smooth"),lower=list(continuous="smooth"))
```


#### Scatter plot matrix with loess smoother {-}
The `ggpairs` function also supports a nonlinear `loess` smoother, but not the
LOWESS smoother implemented in GeoDa. In order to include the latter, it would
be necessary to create a custom function to pass as an argument to the `upper`
and `lower` settings. This is beyond our current scope (again, see 
[the GGobi Github page](http://ggobi.github.io/ggally/rd.html#ggpairs) for techical details). Similarly, if one wanted finer control over the parameters of the
smoothing method (like setting a span), this must be implemented by means
of a custom function.

The loess smoother is passed in the same way as the linear smoother, as an
argument to `continuous = "smooth_loess"`.

```{r}
ggpairs(nyc.data, columns=c("hhsiz00","kids2000", "yrhom02", "pubast00"),
        upper=list(continuous="smooth_loess"),lower=list(continuous="smooth_loess"))
```

The `ggpairs` function has many other customization features to deal with axis
labels, titles, etc., which we do not further pursue here. As mentioned above, just
about anything can be included as a custom function using the **ggplot** API
(for example, using the `wrap` functionality documented on the [Github
pages](http://ggobi.github.io/ggally/rd.html#ggmatrix-helpers)).

Finally, as already alluded to earlier, linking and brushing are not
included in the functionality of **ggplot**. The graphs can be made interactive by means of the
**plotly** package, which we illustrate below for the 3D scatter plot.

## Three Variables: Bubble Chart and 3D Scatter Plot {-}

### Bubble chart {-}
The bubble chart augments the scatter plot with a third dimension, the `size` of
the point (or, bubble). Optionally, a fourth dimension can be added as the `color`
of the point, but this quickly becomes difficult to discern. In GeoDa, the
four dimensions are available by default, with the third and fourth set to the
same variable. In **ggplot**, this is accomplished by setting a third and
potentially fourth aesthetic for `size` and `col` to a variable.

We first illustrate a bubble chart using the variable **kids2000** for the
x-axis, **pubast00** for the y-axis, and **rent2002** for the bubble size. These
are passed as arguments to `aes`. This is followed by the `geom_point` geom.
As before, we can add labels for the x and y axis, as well as a title.

```{r}
ggplot(data=nyc.data,aes(x=kids2000,y=pubast00,size=rent2002)) +
  geom_point() +
  xlab("Percent HH with kids") +
  ylab("Percent HH with public assistance") +
  ggtitle("Bubble Chart") +
  theme(plot.title = element_text(hjust = 0.5))
```

Now, we also introduce the `col` as **rent2002**.

```{r}
ggplot(data=nyc.data,aes(x=kids2000,y=pubast00,size=rent2002,col=rent2002)) +
  geom_point() +
  xlab("Percent HH with kids") +
  ylab("Percent HH with public assistance") +
  ggtitle("Bubble Chart") +
  theme(plot.title = element_text(hjust = 0.5))
```

Note that we used the default color bar for the color argument. There is a wide scope for customization of legends and color schemes in **ggplot**, which is beyond the current scope. In a nutshell, in order to fully mimic the graphs in GeoDa, one would need to use one of the ColorBrewer color schemes, which are available as an option in **ggplot**.

### 3D Scatter Plot {-}

#### Basic 3D scatter plot {-}
The three-dimensional scatter plot is a simple generalization of the two-dimensional
case by creating a graph that projects a 3D cube onto the two-dimensional screen (or paper), as a perspective plot. This is not (currently) supported by **ggplot**, so we resort to the specialized package **scatterplot3d** [see @scatterplot3d03]. This is an older package that predates the layered logic of **ggplot**, and instead uses
the approach taken in the base R `plot` commands.

A rudimentary plot follows from the `scatterplot3d` function to which the variables for the three dimensions are passed as arguments to `x`, `y`, and `z`. In contrast to 
**ggplot**, there is no `data` argument, but the variables must be specified
using the standard `$` notation. We again use the variables **kids2000**,
**pubast00**, and **rent2002**. In addition, we spiff up the graph a bit by
adding a `main` title, as well as titles for `xlab`, `ylab`, and `zlab`. In 
addition, we set the symbol to a filled circle (the default is a hollow circle),
using the base R `pch = 20` argument, and color it red (`color = "red"`), as
in the GeoDa Workbook example. The result is as given below.


```{r}
scatterplot3d(x = nyc.data$kids2000, y = nyc.data$pubast00, z = nyc.data$rent2002,
              main = "NYC 3D Scatterplot",
              xlab = "Percent HH with kids",
              ylab = "Percent HH with public assistance",
              zlab = "Median rent",
              pch = 20,
              color = "red")
```


#### **Plotly** in a nutshell {-}

The static 3D scatter plot is fine as a traditional graph, but is not that
useful for data exploration. To that effect, we will illustrate some functionality
contained in the **plotly** package. This package forms an R interface to
the extensive open source Javascript graphing library of the same name, 
[plotly.js](https://plot.ly/javascript/), which leverages the extensive
collection of D3 charts. The **plotly** package is just one of several
interfaces to **plotly**. Another commonly used one is Plotly.py for Python.

The functionality in **plotly** is huge, and
much more than we can cover here. We refer to the many web resources
for further details. An excellent overview is given in Carson Sievert's
[plotly for R book](https://plotly-book.cpsievert.me), especially in
Chapter 2, [The Plotly Cookbook](https://plotly-book.cpsievert.me/the-plotly-cookbook.html).

There are two main ways to create interactive graphs using **plotly**. One is
to pass the usual arguments to the `plot_ly` command, which has its own
*Grammar of Graphics* syntax. This uses the concept of *traces*, which is similar
to the  *layers* in **ggplot**. The second way uses the `ggplotly` command,
which takes one or more **ggplot** objects and makes them interactive. We will
not cover the second approach, but it is a fairly straightforward way to
make any of our earlier **ggplot** graphs interactive.

The basic arguments to `plot_ly` are the same as for `ggplot`, i.e., the
data set, and the axes (`x`, `y`, and for 3D, `z`). The variable names
are passed in a slightly different way, and use the *formula* notation, with
the variable name prefaced by the `~` symbol. So, for example, if the
x-axis would map to the variable **kids2000**, that would be specified
as `x = ~kids2000`, and similarly for the other axes.

Just like **ggplot**, **plotly** has a layered approach to constructing 
a graph, but instead of using a plus sign to separate the layers, a
*pipe* command, `%>%` is used. Also, the various options for customization
are passed to the respective arguments as a `list`.

We will illustrate the basics of `plot_ly` by constructing a 3D
scatter plot for the same three variables as above.


#### Interacting with the 3D scatter plot  {-}
There are several ways to interact with a plot in **plotly**, but here we will
illustrate some basic functionality to zoom in, zoom out, and rotate the 3D cube,
similar to what is available in GeoDa.

We start with a bare bones graph. We pass the data set and the three variables
to the `plot_ly` function. Since the default *trace* is a scatter plot, this
is all we really need to specify.

```{r}
plot_ly(nyc.data, x = ~kids2000, y = ~pubast00, z = ~rent2002)
```

There are warnings, but it works. Basically, since we did not specify a trace type, the default `scatter3d` is applied, which is exactly what we wanted. Also, since no mode is specified, the default is set to `markers`.

The moment we move the pointer over the graph, a number of small icons appear in the right-hand top. These correspond to different types of interactions that can be carried out. Before we proceed with those, however, we move the pointer to one of the points. The 
*hovering* functionality (here left to the default of listing all data dimensions) will list the values for x, y and z in a small box, and draw the projections to each of the axes.

The second left-most icon at the top of the graph invokes the `Zoom` functionality. Now, moving the pointer back and forth makes the cube smaller or larger. Other interesting options are the `Orbital rotation` and `Turntable rotation`, the two icons
to the right of the home symbol. Both options move the cube around as the pointer changes position. The left-most icon allows a static version of the plot to be downloaded as a png file.

Next, we illustrate a very simple way to add some further information to the graph.
First, we use `add_markers` to turn the observation points red. Note how the
argument `marker` is set equal to a `list` to pass the needed `color` parameter. The `add_markers` command follows the initial `plot_ly` setting after a `%>%`
pipe symbol. A final touch is to set titles for the axes, by means of the
`layout` command and the `scene` option, again after a pipe symbol. The titles are set by means of a `list` command for each of the three `xaxis`, `yaxis`, and `zaxis`.

At this point, when we execute the command, there are no more warnings. Also, we
can interact with the graph in the same way as before.


```{r}
plot_ly(nyc.data, x = ~kids2000, y = ~pubast00, z = ~rent2002) %>%
  add_markers(marker = list(color="red")) %>%
  layout(scene = list(xaxis = list(title = "Percent HH with kids"),
                     yaxis = list(title = "Percent HH with public assistance"),
                     zaxis = list(title = "Median rent")))
```

## True Multivariate EDA: Parallel Coordinate Plot and Conditional Plots {-}
True multivariate EDA deals with situations where more than three variables
are considered. We follow the GeoDa Workbook and illustrate the Parallel Coordinate
Plot, or PCP, and conditional plots. For the former, we again need to resort to
**GGally**, but for the latter, we can exploit the `facet_wrap` and `facet_grid` functions of **ggplot**. In addition, we can turn these plots into interactive graphs by means of the **plotly** functionality.



### Parallel Coordinate Plot (PCP) {-}

#### PCP in **GGally** {-}

The PCP is implemented in the `ggparcoord` function of **GGally**. However, 
its implementation does not follow the regular `columns` specification we
used above for the scatterplot matrix. Instead of passing a list of
variable names, the actual column numbers of the
variables in the data frame must be specified (but those may be in any order).

An easy, though not very elegant way to deal with this is to create a 
subset of the data for those variables to be plotted, and then exploit the
default of `columns = 1:ncol(data)`. In other words, we don't have to
specify the `columns` argument at all.

In the example below, we use the same four variables as in the GeoDa
Workbook: **kids2000**, **rent2002**, **pubast00**, and **yrhom02**.
We first `select` those from the **nyc.data** set to create a subset
we call **pcp.vars**. Then, we pass this subset as the argument to `data`
in `ggparcoord`. The result is a fairly rudimentary PCP, with the axes
organized vertically (in GeoDa, they are horizontal). Many customizations
are possible, for which
detailed options can be found in [the documentation pages](https://www.rdocumentation.org/packages/GGally/versions/1.4.0/topics/ggparcoord).

```{r}
vars <- c("kids2000","rent2002","pubast00","yrhom02")
pcp.vars <- select(nyc.data,vars)
ggparcoord(data = pcp.vars)
```


#### PCP in **plotly** {-}
In **plotly**, the PCP functionality in implemented as the
`type = "parcoords"`. This is passed as the second argument to the `plot_ly`
function (the first argument, as usual, is the data set, **nyc.data**). The
axes of the PCP are specified through the `dimensions` argument. As is the case
in other `plot_ly` examples, they are passed as a list. In this instance, this
is actually a list of lists, one for each axis. In each of these lists, 
we include a `label` for the axis, and a variable as the argument
to the `values` parameter. As before, we need to use a formula format for the
variables and precede their names with the `~` symbol. 


```{r}
plot_ly(nyc.data,type = "parcoords",
          dimensions = list(
            list(label = "Kids", values = ~kids2000),
            list(label = "Public Assistance", values = ~pubast00),
            list(label = "Rent", values = ~rent2002),
            list(label = "Stable", values = ~yrhom02)
            )
          )
```


Once we move the pointer over the graph, a few icons appear on the top right (but fewer than for the 3D scatter plot). The interaction with the graph is not that intuitive,
but once you know what to look for, it is quite powerful. The easiest way to proceed is to click on one of the axes: with the cross hair `+` symbol placed at any location along an axis, clicking will change the color and *select* the observations (lines) covered by the small vertical bar (the pointer will turn into an arrow that points up or down, depending on the direction of the selection). Clicking the cross hair in another position on the same axis turns the selection off.

A second interactive feature allows one to change the order of the axes. For example,
if we move the pointer to the top of the *Stable* axis, it changes from a cross hair to a double sided arrow <->. Pressing down on the pointer now lets us move this axis to the left, e.g., to become the third axis. This can be done even while certain observations are selected. 

Several options for customization of the PCP graph can be found in 
the **plotly** [documentation for `parcoords`](https://plot.ly/r/reference/#parcoords).

### Conditional Plots {-}
Conditional plots are a major feature of the functionality of **ggplot**, where they are referred to as *facetting*, or *small multiples*. This is implemented in the `facet_wrap` and `facet_grid` functions. The main difference between the two approaches is that `facet_grid` is explicitly two-dimensional. In that aspect, it is the closest matches to the conditional plot design in GeoDa.

There is one major difference between the approach taken in GeoDa and that in **ggplot**. In GeoDa, the conditioning variables are typically continuous, and different types of classifications can be applied to them to obtain the actual condition. For example, in the GeoDa Workbook illustration, the variables **hhsiz00** and **yrhom02** are used as conditioning variables for respectively, the x-axis and the y-axis. A classification such as quantiles (e.g., 3 or 2 in the GeoDa Workbook examples) yields the categories for the sub-plots. In **ggplot**, the conditioning is based on a categorical variable that needs to be available in the data set. The facetting formula does not evaluate functions, so the conditioning categories need to be computed beforehand.

There are three so-called helper functions to make this easy: `cut_interval`,
`cut_width`, and `cut_number`. The closest to the median (2 quantiles) conditioning illustrated in the GeoDa Workbook is the `cut_number` function. We pass the variable, e.g., **hhziz00**, and the number of categories, say `n = 2`. This creates the new variable as an R `factor`, giving the intervals that resulted from the cut.

For example, we create a new variable **cut.hhsiz** using a quantile classification with two categories (as in the GeoDa Workbook, the variable will be split on the median
value), by setting `n=2`. We need to use the `$` notation to ensure that the new variable is added to the relevant data set. Since we only have 55 observations, we can easily list the
full set of values to verify. Internally, they are stored as *factors* (hence,
the summary of the `Levels` at the end of the listing).

```{r}
nyc.data$cut.hhsiz <- cut_number(nyc.data$hhsiz00,n=2)
nyc.data$cut.hhsiz
```

And, similarly for **cut.yrhom**:

```{r}
nyc.data$cut.yrhom <- cut_number(nyc.data$yrhom02,n=2)
nyc.data$cut.yrhom
```

If we compare the breakpoints to the ones in Figure 34 of the Workbook, we see
that they are close, but not exactly the same, i.e., 2.72 vs. 2.703 in GeoDa, and
12.4 vs. 12.368. More
precisely, upon closer examination, we find that for **cut.hhsiz**, the lower
group has 28 observations vs. 27 in GeoDa. Since we have so few data points,
this may lead to slight differences in the graphs.

At this point, we can set up the conditioning in the `facet_grid` function,
expressed as a formula, with the row conditioning variable first. Note that
the row conditioning variable is the y-axis in GeoDa, and the column
conditioning variable is the x-axis.
For example, with our new categories **cut.hhsiz** and **cut.yrhom**, this would be
`facet_grid(cut.yrhom ~ cut.hhsiz)`. 

One final aspect is how the categories are ordered in the graph. The 
default (`as.table=TRUE`) is to have the highest category in the lower-right
corner. In order to mimic the organization in GeoDa, we set `as.table=FALSE`.
This results in the highest category being in the upper-right corner.

We now illustrate this for a conditional scatter plot and a conditional histogram.

#### Conditional scatter plot {-}
We replicate the example in the GeoDa Workbook and condition a scatter plot
with **kids2000** on the x-axis and **pubast00** on the y-axis. We set these
two variables as `x` and `y` in the `aes` argument of `ggplot`. Next, we 
specify the geom as `geom_point`, for the default scatter plot. Finally, we
add the `facet_grid` command.

```{r}
ggplot(data=nyc.data,aes(x=kids2000,y=pubast00)) +
  geom_point() +
  facet_grid(cut.yrhom ~ cut.hhsiz,as.table=FALSE)
```

We can add a linear smoother by means of `geom_smooth(method="lm")`:

```{r}
ggplot(data=nyc.data,aes(x=kids2000,y=pubast00)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_grid(cut.yrhom ~ cut.hhsiz,as.table=FALSE)
```

We can also add a loess smoother by means of `geom_smooth(method="loess")`:

```{r}
ggplot(data=nyc.data,aes(x=kids2000,y=pubast00)) +
  geom_point() +
  geom_smooth(method="loess") +
  facet_grid(cut.yrhom ~ cut.hhsiz,as.table=FALSE)
```

As is the case for all graphs in **ggplot**, many further customizations can
be added, but we do not consider that further.

#### Conditional histogram {-}
We conclude with a conditional histogram for the variable **pubast00**.
The principle is the same as before. The only difference is that now
only one variable needs to be specified in `aes`, and the `geom_histogram`
is used. As we did earlier, we set the `bins=7` (the default of 30 is not
appropriate in this example). The resulting graph differs slightly from the
example in the GeoDa Workbook due to different bin widths. With some
customization, they can be made to look exactly the same, but we won't
pursue that here.

```{r}
ggplot(data=nyc.data,aes(pubast00)) +
  geom_histogram(bins=7) +
  facet_grid(cut.yrhom ~ cut.hhsiz,as.table=FALSE)
```


<br>

## References {-}



