---
title: "Contiguity-Based Spatial Weights"
subtitle: "R Notes"
author: "Luc Anselin and Grant Morrison^[University of Chicago, Center for Spatial Data Science -- anselin@uchicago.edu,morrisonge@uchicago.edu]"
date: "08/08/2018"
output:
  html_document:
    fig_caption: yes
    self_contained: no
    toc: yes
    toc_depth: 4
    css: tutor.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

## Introduction {-}

This notebook cover the functionality of the [Contiguity-Based Spatial Weights](https://geodacenter.github.io/workbook/4a_contig_weights/lab4a.html) section of the GeoDa workbook. We refer to that document for details on the methodology, references, etc. The goal of these notes is to approximate as closely as possible the operations carried out using GeoDa by means of a range of R packages.

The notes are written with R beginners in mind, more seasoned R users can probably skip most of the comments
on data structures and other R particulars. Also, as always in R, there are typically several ways to achieve a specific objective, so what is shown here is just one way that works, but there often are others (that may even be more elegant, work faster, or scale better).

For this notebook, we use U.S. Homocide data. Our goal in this lab is show how to implement contiguity based spatial weights 


```{r}

```
### Objectives {-}

After completing the notebook, you should know how to carry out the following tasks:

- Construct rook and queen contiguity-based spatial weights

- Compute higher order contiguity weights

- Save weights information

- Assess the characteristics of spatial weights

- Visualize the graph structure of spatial weights

- Identify the neighbors of selected observations

#### R Packages used {-}

- **sf**: To read in the shapefile, add centroids, and create the neighbors lists

- **purrr**: Used to map a function over each element of a vector

- **ggplot2**: To make a connectivity histogram

- **spdep**: Save weights files and create neighbors lists of higher order

- **geodaData: Load the data for the notebook.

#### R Commands used {-}

Below follows a list of the commands used in this notebook. For further details
and a comprehensive list of options, please consult the 
[R documentation](https://www.rdocumentation.org).

- **Base R**: `install.packages`, `library`, `setwd`, `class`, `str`, `lapply`, `attributes`, `summary`, `head`, `seq`

- **sf**: `st_read`, `plot`, `st_centroid`, `st_relate`

- **purrr**: `map_dbl`

- **ggplot2**: `ggplot`, `geom_histogram`, `aes`, `xlab`

- **spdep**: `write.nb.gal`, `nblag`, `nblag_cumul`, `card`

## Preliminaries {-}

Before starting, make sure to have the latest version of R and of packages that are compiled for the matching version of R (this document was created using R 3.5.1 of 2018-07-02). Also, optionally, set a working directory, even though we will not
actually be saving any files.^[Use `setwd(directorypath)` to specify the working directory.]

### Load packages {-}

First, we load all the required packages using the `library` command. If you don't have some of these in your system, make sure to install them first as well as
their dependencies.^[Use 
`install.packages(packagename)`.] You will get an error message if something is missing. If needed, just install the missing piece and everything will work after that.


```{r}
library(sf)
library(spdep)
library(purrr)
library(ggplot2)
library(geodaData)
```


### geodaData {-}

All of the data for the R notebooks is available in the **geodaData**
package. We loaded the library earlier, now to access the individual
data sets, we use the double colon notation. This works similar to
to accessing a variable with `$`, in that a drop down menu will 
appear with a list of the datasets included in the package. For this
notebook, we use `ncovr`.


```{r}
us.bound  <- geodaData::ncovr
```

## Contiguity Weights {-}

Contiguity means that two spatial units share a common border of non-zero length.
Operationally, we can further distinguish between a rook and a queen criterion of
contiguity, in analogy to the moves allowed for the such-named pieces on a chess
board.

The rook criterion defines neighbors by the existence of a common edge between two
spatial units. The queen criterion is somewhat more encompassing and defines
neighbors as spatial units sharing a common edge or a common vertex.4 Therefore, the
number of neighbors according to the queen criterion will always be at least as
large as for the rook criterion.

In practice, the construction of the spatial weights from the geometry of the data
cannot be done by visual inspection or manual calculation, except in the most
trivial of situations. To assess whether two polygons are contiguous requires the
use of explicit spatial data structures to deal with the location and arrangement of
the polygons. This is implemented through the spatial weights functionality in
GeoDa. We will do this with **sf** and **spdep** libraries.

We will create our neighbors using **sf** first, as the **spdep** library doesn't 
allow us to create neighbors lists directy as with the older **sp** library. 
When we create neighbors in **sf** we will get a class of **sgbp**(sparse geometry binary predicate), which is similar
to the standard **nb** class used in **spdep**, but is not quite compatible. We will
have to convert from **sgbp** to **nb**, which is not too difficult, as the classes are
very similar.

It is important to keep in mind that the spatial weights are critically dependent on
the quality of the spatial data source (GIS) from which they are constructed.
Problems with the topology in the GIS (e.g., slivers) will result in inaccuracies
for the neighbor relations included in the spatial weights. In practice, it is
essential to check the characteristics of the weights for any evidence of problems.
When problems are detected, the solution is to go back to the GIS and fix or clean
the topology of the data set. Editing of spatial layers is not implemented in GeoDa,
but this is a routine operation in most GIS software.



### Rook Contiguity {-}

We first start with rook contiguity, which are neighbors that are connected by a
common side. To create our neighbors list we start by making a 
function that will do the rook contiguity. To do this we start with the `function`
command and use `a, a = b` for our inputs. We do this reduce unnecessary typing,
as the two parameters are both going to be our **sf** object. From here we just
need `st_relate`. `st_relate` computes relationships between pairs of geometries,
or matches it to a given pattern. This function also has a parameter for a specified 
pattern. This pattern will refer to a DE9-IM relationship between x[i] and y[j]. We
don't need to go into detail on this to utilize rook contiguity, but you can check
out the basics of DE9-IM at [DE9-IM](https://en.wikipedia.org/wiki/DE-9IM). All we
need for rook contiguity is the pattern input `"F***1****"`. This gives us the
correct DE9-IM relationship for rook contiguity. For more documentation on
`st_relate` check out [st_relate documentation](https://www.rdocumentation.org/packages/sf/versions/0.6-3/topics/st_relate)

```{r}
st_rook = function(a, b = a) st_relate(a, b, pattern = "F***1****")
```


```{r}
sf.sgbp.rook <- st_rook(us.bound)
```




We now check the class of our resulting object from the the `st_rook` function
with the base R function `class`. It is **sgbp**, which we will have to convert in order to
work with **spdep** and the `plot` function later on.

```{r}
class(sf.sgbp.rook)
```


We can get a comprehensive view of what the **sgbp** data structure looks like with
the `str` command. We will see a list of 3085 vectors that contain the id numbers of
each observation's neighbors. Along with this we will see some attributes. These
are the same attributes as an **nb** and the data structure follows the same form
as **nb**. This is important for later operations where we will need to convert
to type **nb**.


```{r}
str(sf.sgbp.rook)
```

Now we will start our conversion from class **sgbp** to **nb**. To do this,
we need to change the class explicitly, and take the precaution to represent 
observations with no neighbors with the integer 0. Our data set doesn't have
any observations without neighbors, but in ones with these, it will mess everything
up, if not dealt with. 

We start with the `function` operator. The input for our function will be an object
of class **sgbp**, denoted by **x**. We store the attributes in **attrs**, as we will
need to reapply them later. Now we deal we observation with no neighbors. We will
use `lapply`, which applies a function to each element of a list or vector. As for the
input function, we make one that checks the length of an element of our list for
0(meaning no neighbors) and returns 0 if the element is empty and the element
otherwise. This can be a bit confusing, for more information on `lapply`, check out 
[lapply documentaion](https://www.rdocumentation.org/packages/base/versions/3.5.1/topics/lapply)

From here we will have dealt with observation with no neighbors, but will need to 
reapply our attributes to the resulting structure from the `lapply` function. This 
is done by calling the `attributes` function and assigning our stored attributes.
The final step is to explictly change the class to **nb** by using the `class` function
and assigning `"nb"`. We then return our object **x**.



```{r}
as.nb.sgbp <- function(x, ...) {
  attrs <- attributes(x)
  x <- lapply(x, function(i) { if(length(i) == 0L) 0L else i } )
  attributes(x) <- attrs
  class(x) <- "nb"
  x
}
```


Now we use the function we created to convert our **sgbp** object to **nb**.

```{r}
sf.nb.rook <- as.nb.sgbp(sf.sgbp.rook)
```



Now we take a comprehensive look at our resulting **nb** object with the `str` command.
It is essentially the same as before, just the class now says **nb** instead of **sgbp**. This makes our earlier steps seem a bit unnecesssary and overcomplicated, 
but they are important, as all datasets, will not have the same circumstances as 
this one, and it is important for these things to work across the board.

```{r}
str(sf.nb.rook)
```

Now that we are converted to **nb**, we can use the `summary` command to give us more
useful information about the neighbours list.

```{r}
summary(sf.nb.rook)
```

We use the `class` command to check and make sure we have **nb**.

```{r}
class(sf.nb.rook)
```

We check the length to make sure it corresponds with the GeoDa tutorial example.

```{r}
length(sf.nb.rook)
```




### Queen Contiguity {-}


We proceed in the same fashion to construct queen contiguity weights. The difference
between the rook and queen criterion to determine neighbors is that the latter also
includes common vertices. This makes the greatest difference for regular grids (square
polygons), where the rook criterion will result in four neighbors (except for edge
cases) and the queen criterion will yield eight. For irregular polygons (like most
areal units encountered in practice), the differences will be slight. In order to deal
with potential inaccuracies in the polygon file (such as rounding errors), using the
queen criterion is recommended in practice. Hence it is also the default for
contiguity weights.

To make our queen contiguity weights, we make a function using `st_relate` and 
specifying a DE9-IM pattern. For queen contiguity, our pattern is `"F***T****"`. We
we don't really need to go into why the pattern is what it is for our purposes, but
[DE9-IM](https://en.wikipedia.org/wiki/DE-9IM) and [st_relate documentation](https://www.rdocumentation.org/packages/sf/versions/0.6-3/topics/st_relate) can help explain this to a degree.


```{r}
st_queen <- function(a, b = a) st_relate(a, b, pattern = "F***T****")
```


Now we use our `st_queen` function to get another **sgbp** neighbour list

```{r}
sf.sgbp.queen <- st_queen(us.bound)
```

To convert to type **nb** we use `as.nb.sgbp`, which we created earlier.


```{r}
sf.nb.queen <- as.nb.sgbp(sf.sgbp.queen)
```


```{r}
str(sf.nb.queen)
```


## Higher Order Contiguity {-}

Now we move on to higher order contiguity weights. To make these we will need the
**spdep** package. We will use the `nblag` and `nblag_cumul` functions to compute
the higher order weights.

The `nblag` function takes a neighbor list of class **nb** and an order as parameters.
It will give us a list of neighbors lists. One for 1st order neighbors and one for
second order neighbors. We can select from this data structure by double bracket
notation, for instance `[[1]]` will give the **nb** object for first order neighbors.
Using a `2` instead of a `1` will give us the second order neighbors.


```{r}
second.order.queen <- nblag(sf.nb.queen, 2)
```

We first take a comprehensive look at **second.order.queen** to see the resulting
data structure from `nblag`. As said earlier, it is a list of 2 **nb** objects. One
for 1st order neighbors and one for 2nd order. It is important to examine this, so
we can make use of specific elements of the data structure in our visualizations
of the neighbors.

```{r}
str(second.order.queen)
```

Now we can look at, specifically, the 2nd order neighbors by using the double bracket
selection.

```{r}
str(second.order.queen[[2]])
```

We store the second order neighbours in **sf.nb.queen2** to make things simpler
for our visuals.

```{r}
sf.nb.queen2 <- second.order.queen[[2]]
```




Now if we want to include both 1st and 2nd order neighbors in our visualizations, we
need the `nblag_cumul` function. This will give us an **nb** object with both 1st and 2nd order neighbors, instead of separating them as in the result of `nblag`. To use this
function, we need the resulting object from the `nblag` function. `nblag_cumul` combines the two sepatae **nb** objects from the `nblag` function into one.


```{r}
second.order.queen.cumul <- nblag_cumul(second.order.queen)
```


Here we take another in depth look at result. We can see that, we now have one **nb**
object with both 1st and 2nd order neighbors.

```{r}
str(second.order.queen.cumul)
```



## Visualizing Contiguity Neighbors {-}



### Connectivity Graph {-}

A connectivity graph takes a point and displays a line to each neighboring point. We are
working with polygons at the moment, so we will need to get points in order to make
our connectivity graphs. The most typically method for this will be polygon centroids.
We will calculate these in the **sf** package before moving onto the graphs.



#### Getting Latitude and Longitude of Polygon Centroids {-}

We will need points to associate with each polygon before we can make our connectivity
graph. It will be a little more complicated than just running `st_centroid` on
the **sf** object: **us.bound**. We need the coordinates in a separate data frame
for this to work. To do this we will use a mapping function. The mapping function
applies a given function to each element of a vector and returns a vector of the 
same length. Our input vector will be the geometry column of **us.bound**. Our
function will be `st_centroid`. We will be using `map_dbl` variation of `map` from
the **purrr** package. For more documentation, check out [map documentation](https://www.rdocumentation.org/packages/purrr/versions/0.2.5/topics/map)


To get our longitude values we map the `st_centroid` function over the geometry 
column of **us.bound** and access the longitude value through double bracket notation
[[]] and 1. This allows us to get only the longitude, which is the first value in each
centroid.


```{r}
longitude <- map_dbl(us.bound$geometry, ~st_centroid(.x)[[1]])
```

We do the same for latitude with one key difference. We access the second value per
each centroid with `[[2]]`.

```{r}
latitude <- map_dbl(us.bound$geometry, ~st_centroid(.x)[[2]])
```

Now that we have latitude and longitude, we use `cbind` to put longitude and latitude
into the same object.

```{r}
coords <- cbind(longitude, latitude)
```


We check the first few observations to see if things are formatted correctly.
```{r}
head(coords)
```




#### Rook Connectivity Graphs {-}

Now that we have coordinates for each of our observations, we can plot the neighbours
through a connectivity graph. We just need the basic `plot` function and to enter
our neighbour list as the first parameter, then our coordinates as the second. We
add extra customizations, so we can actually see what is going on in the graph.
We use `lwd = .2` to make the line length less thick, we set the `col ="blue"`, and 
finally `cex = .5` to make the point symbols smaller. The default options make the
graph near indecipherable, so it is important to add these extra parameters.

```{r}
plot(sf.nb.rook, coords, lwd=.2, col="blue", cex = .5)
```



#### Queen Connectivity Graphs {-}

Now we will do the same for queen contiguity. The queen variation will have more lines,
as there are more neighbors due to shared vertices counting in the neighbors
relationship. It is hard to spot the differences between the two plots, as they are
both visually busy 

##### First Order {-}

The same form is followed for queen contiguity as we are working with the same class:
a **nb** object and a corresponding list of coordinates.

```{r}
plot(sf.nb.queen, coords, lwd=.2, col="blue", cex = .5)
```

##### Second Order {-}

Here we take a look at a connectivity graph for second order neighbors. All we need
is the second order **nb** object that we created earlier and our coordinates. This
graph will look significantly different from the ones we've made so far, as second 
order tends to have more neighbors than first order.

```{r}
plot(sf.nb.queen2, coords, lwd=.2, col="blue", cex = .5)
```


##### Cumulative {-}

We will take a look a plot with both second and first order neighbors here. It will
be a very busy graph. 


```{r}
plot(second.order.queen.cumul, coords, lwd=.2, col="blue", cex = .5)
```




### Connectivity Histogram {-}

A connectivity histogram shows the number of observations for each 
cardinality of neighbors(how many observations have the same number of neighbors).
We first have to get the cardianlity before we can make the histograms. We can
do this by using the `card` function from the **spdep** package. This will give
us a vector of 3085 observations with each one containing the corresponding
number of neighbors from the input neighbors list.



```{r}
rook.card <- card(sf.nb.rook)
```


We take a look at the first few observation with the `head` function and as we can see,
it is a vector with number of neighbors for each observation of the **nb** object.
```{r}

head(rook.card)
```

Now we check the length and it is what we expected.
```{r}
length(rook.card)
```

We can visualize the cardinality through a **ggplot2** histogram. For this we just
need to call the `ggplot` function and add `geom_histogram` as a layer. We specify the
`aes()` in `geom_histogram` to be our rook cardinality. This will give us a basic plot, which
we will customize further a little later.

```{r}
ggplot() +
  geom_histogram(aes(x=rook.card))
```



Here we will add a few customizations to make the plot nicer. The most important is with
the `breaks =` parameter, as the bars are spaced out kind of weirdly in our opening plot.
We can avoid manuelly typing in the breaks by using the `seq` function, which will give us
a list of numbers 0 to 13, incrementing by 1 each time. We specify the range in the first
two arguments and then use `by =` to pick the number to increment by. In our case, 1 makes
the most sense as you cannot have half of a neighbor.

```{r}
ggplot() +
  geom_histogram(aes(x=rook.card), breaks = seq(0,13, by = 1)) +
  xlab("number of neighbours")
```


With the `summary` command we can get a look at the summary
statistics for the cardinality of the neighbors list.
```{r}
summary(rook.card)
```




## Saving Neighbors {-}
To save our neighbors list, we use the `write.nb.gal` function from
the **spdep** package. The file format is a GAL Lattice file. We
input our neighbors list, and the the filename second. We have two 
options from this point. We can save the file with the old style
or the new GeoDa header style.

### Oldstyle {-}

The oldstyle just saves the neighbor list with then number of observations
at the top of the file. We can save in this format by setting `oldstyle =`
to `TRUE`.

```{r}
write.nb.gal(sf.nb.rook, "rook_contiguity_old.gal", oldstyle = TRUE)
```



### GeoDA Header Format {-}

The new GeoDA header style, also takes the shapefile name taken from GAL file for
the dataset and the region id indicator variable name. It puts this information
in a header for the file. All we have to do is set `oldstyle =` to FALSE and
enter names for `shpfile` and `ind` parameters.

```{r}
write.nb.gal(sf.nb.rook, "rook_contiguity_new.gal", oldstyle = FALSE,shpfile ="NAT.shp", ind ="region id")
```
















