---
title: "Distance-Based Spatial Weights"
subtitle: "R Notes"
author: "Luc Anselin and Grant Morrison^[University of Chicago, Center for Spatial Data Science -- anselin@uchicago.edu,morrisonge@uchicago.edu]"
date: "11/17/2019"
output:
  html_document:
    fig_caption: yes
    self_contained: no
    toc: yes
    toc_depth: 4
    css: tutor.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

## Introduction {-}

This notebook cover the functionality of the [Distance-Based Spatial Weights](https://geodacenter.github.io/workbook/4b_dist_weights/lab4b.html) section of the GeoDa workbook. We refer to that document for details on the methodology, references, etc. The goal of these notes is to approximate as closely as possible the operations carried out using GeoDa by means of a range of R packages.

The notes are written with R beginners in mind, more seasoned R users can probably skip most of the comments
on data structures and other R particulars. Also, as always in R, there are typically several ways to achieve a specific objective, so what is shown here is just one way that works, but there often are others (that may even be more elegant, work faster, or scale better).

For this notebook, we use Cleveland homesale point data. Our goal in this lab is show how to implement distance-band spatial weights 



### Objectives {-}

After completing the notebook, you should know how to carry out the following tasks:

- Construct distance band spatial weights

- Assess the characteristics of distance-based weights

- Assess the effect of the max-min distance cut-off

- Identify isolates

- Construct k-nearest neighbor spatial weights

- Create Thiessen polygons from a point layer

- Construct contiguity weights for points and distance weights for polygons

- Understand the use of great circle distance

#### R Packages used {-}

- **tmap**: To plot our points on a base map

- **sf**: Used to read the shapefiles in, make contiguity weights,
and convert from **sp**, 

- **spdep**: Used to create distance neighbors and contiguity neighbors

- **ggplot2**: To make connectivity histograms

- **deldir**: To make Thiessen polygons.

- **sp**: Used to get an intermediate data structure to get the Thiessen polygons to **sf** class

_ **purr**:  Used for a mapping function

- **geodaData**: To get the data for this notebook


#### R Commands used {-}

Below follows a list of the commands used in this notebook. For further details
and a comprehensive list of options, please consult the 
[R documentation](https://www.rdocumentation.org).

- **Base R**: `install.packages`, `library`, `setwd`, `head`, `str`, `summary`, `class`, `cbind`, `unlist`, `max`, `attributes`, `class`, `list`, `lapply`, `rbind`,
`sapply`, `length`, `seq`, `as.character`, `vector`, `data.frame`

- **tmap**: `tm_shape`, `tmap_mode`, `tm_dots`

- **sf**: `st_read`, `plot`, `st_as_sf`, `st_relate`

- **spdep**: `knn2nb`, `knearneigh`, `dnearneigh`, `nbdists`, `card`

- **ggplot2**: `ggplot`, `geom_histogram`, `xlab`

- **deldir**: `deldir`, `tile_list`

- **sp**: `Polygon`, `Polygons`, `SpatialPolygons`, `SpatialPolygonsDataFrame`

_ **purr**: `map_dbl`

## Preliminaries {-}

Before starting, make sure to have the latest version of R and of packages that are compiled for the matching version of R (this document was created using R 3.5.1 of 2018-07-02). Also, optionally, set a working directory, even though we will not
actually be saving any files.^[Use `setwd(directorypath)` to specify the working directory.]

### Load packages {-}

First, we load all the required packages using the `library` command. If you don't have some of these in your system, make sure to install them first as well as
their dependencies.^[Use 
`install.packages(packagename)`.] You will get an error message if something is missing. If needed, just install the missing piece and everything will work after that.


```{r}
library(tmap)
library(sf)
library(spdep)
library(ggplot2)
library(deldir)
library(sp)
library(purrr)
library(geodaData)
```

### geodaData {-}


All of the data for the R notebooks is available in the **geodaData**
package. We loaded the library earlier, now to access the individual
data sets, we use the double colon notation. This works similar to
to accessing a variable with `$`, in that a drop down menu will 
appear with a list of the datasets included in the package. For this
notebook, we use `ncovr`.

```{r}
clev.points <- geodaData::clev_pts
```


```{r}
plot(clev.points)
```



## Visualizing point data {-}

To get a cursory look at the point data with a basemap, we will use **tmap**. To get 
this done we first have to switch from mode `plot` to mode `view` with the function
`tmap_mode`. From there we use `tm_shape` and `tm_dots` to display our points on the
basemap. The only argument we need to pass is the shapefile into the `tm_shape` function.

```{r}
tmap_mode("view")
tm_shape(clev.points) +
  tm_dots()
```

Now we switch back to `plot` mode for the rest of the notebook with `tmap_mode("plot")`.
```{r}
tmap_mode("plot")
```

## Distance-Band Weights {-}


### Concepts {-}


#### Distance Metric {-}

The core input into the determination of a neighbor relation for distance-based spatial weights is a formal measure of distance, or a distance metric. The most familiar special case is the Euclidean or straight line distance, $d_{ij}$, as the crow flies:
$$ d_{ij} = \sqrt{(x_i-x_j)^2 + (y_i - y_j)^2}$$

for two points i and j with respective coordinates $(x_i,y_i)$ and $(x_j,y_j)$. 


#### Great Circle distance {-}


Euclidean inter-point distances are only meaningful when the coordinates are recorded
on a plane, i.e., for projected points.

In practice, one often works with unprojected points, expressed as degrees of latitude
and longitude, in which case using a straight line distance measure is inappropriate,
since it ignores the curvature of the earth. This is especially the case for longer
distances, such as from the East Coast to the West Coast in the U.S.

The proper distance measure in this case is the so-called arc distance or great circle
distance. This takes the latitude and longitude in decimal degrees as input into a
conversion formula.3 Decimal degrees are obtained from the degree-minute-second value
as degrees + minutes/60 + seconds/3600.

The latitude and longitude in decimal degrees are converted into radians as:

$$Lat_r = (Lat_d - 90) * \pi/180$$
$$Lon_r = Lon_d * \pi/180$$

where the subscripts d and r refer respectively to decimal degrees and radians, and
π=3.14159… With $\Delta Lon = Lon_r(j)- Lon_r(i)$, the expression for the arc distance
is:

$$d_{ij} = R * arccos[cos(\Delta Lon) * sin(Lat_{r(i)}) * sin(Lat_{r(j)}) + cos(Lat_{r(i)}) * cos(Lat_{r(j)})]$$
or equivalently:

$$d_{ij} = R * arccos[cos(\Delta Lon) * cos(Lat_{r(i)}) * cos(Lat_{r(j)}) + sin(Lat_{r(i)}) * sin(Lat_{r(j)})]$$

where R is the radius of the earth. In GeoDa, the arc distance is obtained in miles with R = 3959, and in kilometers with R = 6371.

These calculated distance values are only approximate, since the radius of the earth is taken
at the equator. A more precise measure would take into account the actual latitude at which
the distance is measured. In addition, the earth’s shape is much more complex than a sphere,
but the approximation serves our purposes.

#### Distance-band weights {-}

The most straightforward spatial weights matrix constructed from a distance measure is
obtained when i and j are considered neighbors whenever j falls within a critical distance
band from i. More precisely, $w_{ij}=1$ when $d_{ij} \leq \delta $ , and $w_{ij}=0$ otherwise,
where $\delta$ is a preset critical distance cutoff.

In order to avoid isolates (islands) that would result from too stringent a critical distance,
the distance must be chosen such that each location has at least one neighbor. Such a distance
conforms to a max-min criterion, i.e., it is the largest of the nearest neighbor distances.

In practice, the max-min criterion often leads to too many neighbors for locations that are
somewhat clustered, since the critical distance is determined by the points that are
furthest apart. This problem frequently occurs when the density of the points is uneven
across the data set, such as when some of the points are clustered and others more spread
out. We revisit this problem in the illustrations below.


### Creating distance-band weights {-}


In order to start the distance based neighbors, we first need to compute a threshold
value(The minimum distance that gives each point at least one neighbor). We can
find this using k-nearest neighbors, which will be covered in more depth later
in the notebook. We find the list of k-nearest neighbors for **k = 1** then find the
max distance between two neighbors in this list, which we then use as the upper 
distance parameter in the `dnearneigh`.

#### Computing Critical Threshold {-}

Before we move forward, we need to x and y coordinates in a matrix. This is easily
done by using `cbind` on the x and y coordinate columns of **clev.points**. `cbind`
puts together vectors as columns in a matrix, so it is perfect for this task.

```{r}
coords <- cbind(clev.points$x,clev.points$y)
```

To find the critical threshold, we first find the k-nearest neighbors for k = 1. This
will give us a list, where each point has exactly one neighbor. To do this, we
use the 'knearneigh` function from **spdep** library. This will give a class of
**knn**, which is similar to class **nb**, but will need to be converted anyways.

```{r}
knn1 <- knearneigh(coords)
```

Here we take a comprehensive look at our resulting **knn** object with the `str` command.
```{r}
str(knn1)
```

Now we convert to **nb**. This simple, as there is a built in function for it in
the **spdep** library: `knn2nb`.
```{r}
k1 <- knn2nb(knn1)
```


Computing the critical threshold will require a few functions, now that we have a neighbors
list. First step is to get the distances between each point and it's closest neighbor.
This can be done with the `nbdists`. With these distances, we just need to find the 
maximum. For this we use the `max` command. However, we cannot do this with lists, so we must 
first get a data type that works for the `max` command, in our case, we use `unlist`

```{r}
critical.threshold <- max(unlist(nbdists(k1,coords)))
critical.threshold
```

#### Computing distance-band weights {-}

With our critical threshold, we have a baseline value to work with for our distance-band
neighbors. To make this neighbor's list we use `dnearneigh` from the **spdep** package.
The parameters necessary are the coordinates, the lower distance bound, and the upper
distance bound. We enter these in the the above order. Another important parameter is
the `longlat`. This is used for point data in longitude and latitude form. It is
necessary to use this to get great circle distance instead of euclidean for accuracy
purposes.

```{r}
nb.dist.band <- dnearneigh(coords, 0, critical.threshold)
```

### Weights characteristics {-}

We can examine the characteristics of these weights through a connectivity graph, 
connectivity histogram, and summary statistics.


#### Weights summary {-}

We can use the base R `summary` command to get a comprehensive look at our distance-band
weights object. This will give us a lot of useful information, that we cannot get with
the `str` command.
```{r}
summary(nb.dist.band)
```


#### Connectivity histogram {-}

Our method for making connectivity histograms will be the same as for contiguity
based weights. We first need to get the cardinality for each observation(the number of neighbors for each observation). This is done by the `card` function from the 
**spdep** library. Our result from this function will be a vector of the number
of neighbors for each location.

```{r}
dist.band.card <- card(nb.dist.band)
dist.band.card
```

Once, we have our cardinality, we can make a histogram to see the distribution
of the number of neighbors. we will do this with **ggplot2**, though base R also has a 
built in histogram function. To make our histogram, we start with the `ggplot` function,
then add a histogram layer. We add the additional layer with the `+` operator. Next,
we use `geom_histogram` to add our histogram. The only parameter we need is an `aes`.
We use the `aes` function and enter our cardinality(**dist.band.card**) as the variable
for the x-axis. Beyond this, we change the x-axis label to be more informative with 
the `xlab` function.

```{r}
ggplot() +
  geom_histogram(aes(x=dist.band.card)) +
  xlab("Number of Neighbors")
```


#### Connectivity graph {-}

For our connectivity graph, we follow the same steps as for the contiguity weights.
We use the `plot` command and use our neighbors list and coordinates as inputs. We
use additional parameters to make the graph easier to look at. We shorten the line
width with `lwd =.2` and make the points smaller with `cex = .5`. The default points
are much to big, and it is hard to get sense of the graphs structure.

```{r}
plot(nb.dist.band, coords, lwd=.2, col="blue", cex = .5)
```

We can see that we have no isolates from our graph, as we expect from using a critical
threshold approach. Our graphs structure consists of two subgraphs and two pairs 
of points.


### Isolates {-}

So far, we have used the default cut-off value for the distance band. However, the
function is flexible enough that we can type in any value for the cut-off, or use the
movable button to drag to any value larger than the minimum. Sometimes, theoretical
or policy considerations suggest a specific value for the cut-off that may be smaller
than the max-min distance.

From our example we know the critical threshold is 3598, so we can pick a value lower
to get some isolates in our distance-band weights. We will use 1500 ft, same as the
corresponding GeoDa workbook example.

The only change we need to make to do this, is make the upper distance bound 1500,
instead of **critical.threshold**.

```{r}
dist.band.iso <- dnearneigh(coords, 0, 1500)
```


#### Isolates in the connectivity histogram {-}

We can get a measure of the number of isolates by looking at the left most bar
of a connectivity histogram. To make this, we follow the same procedure as the
earlier connectivity histograms. We first get the cardinality, then use **ggplot2**
to make the histogram.


```{r}
iso.card <- card(dist.band.iso)
```


```{r}
ggplot() +
  geom_histogram(aes(x=iso.card)) +
  xlab("Number of Neighbors")
```
Our resulting histogram has 24 isolates, and is much more compact than the original
one with the critical threshold as the upper distance bound in our distance-band
weights.



#### Isolates in the connectivity graph {-}

The most dramatic visualization of the isolates is given by the connectivity graph. The 24 points without an edge to another point are easily identified

```{r}
plot(dist.band.iso, coords, lwd=.2, col="blue", cex = .5)
```


#### How to deal with isolates {-}


Since the isolated observations are not included in the spatial weights (in effect,
the corresponding row in the spatial weights matrix consists of zeros), they are not
accounted for in any spatial analysis, such as tests for spatial autocorrelation, or
spatial regression. For all practical purposes, they should be removed from such
analysis. However, they are fine to be included in a traditional non-spatial data
analysis.

Ignoring isolates may cause problems in the calculation of spatially lagged variables,
or measures of local spatial autocorrelation. By construction, the spatially lagged
variable will be zero, which may suggest spurious correlations.

Alternatives where isolates are avoided by design are the K-nearest neighbor weights
and contiguity weights constructed from the Thiessen polygons for the points. They are
discussed next.




## K-Nearest Neighbor Weights {-}

### Concept {-}


As mentioned, an alternative type of distance-based spatial weights that avoids the
problem of isolates are k-nearest neighbor weights. In contrast to the distance band,
this is not a symmetric relation. The fact that B is the nearest neighbor to A does
not imply that A is the nearest neighbor to B. There may be another point C that is
actually closer to B than A. This asymmetry can cause problems in analysis that depend
on the intrinsic symmetry of the weights (e.g., some algorithms to estimate spatial
regression models). One solution is to replace the original weights matrix **W** by
(**W**+**W′**)/2, which is symmetric by construction. GeoDa currently does not
implement this approach.

A potential issue with k-nearest neighbor weights is the occurrence of ties, i.e.,
when more than one location j has the same distance from i. A number of solutions
exist to break the tie, from randomly selecting one of the k-th order neighbors, to
including all of them. In GeoDa, random selection is implemented.


### Creating KNN Weights {-}

To create our KNN weights, we need two functions from the **spdep** library:
`knearneigh` and `knn2nb`. We first use `knearneigh` to get a class of **knn**,
as we did earlier to find the critical threshold. This time we assign `k =` a value
of 6. This means each observation will get a list of the 6 closest points. We then
use `knn2nb` to convert from class **knn** to class **nb**. 


```{r}
k6 <- knn2nb(knearneigh(coords, k = 6))
```



### Properties of KNN weights {-}

One drawback of the k-nearest neighbor approach is that it ignores the distances
involved. The first k neighbors are selected, irrespective of how near or how far they
may be. This suggests a notion of distance decay that is not absolute, but relative,
in the sense of intervening opportunities (e.g., you consider the two closest grocery
stores, irrespective of how far they may be).


Again, we can also use the connectivity histogram and the connectivity map to inspect the neighbor characteristics of the observations. However, in this case, the histogram doesn’t make much sense, since all observations have the same number of neighbors, as
shown by our histogram.

```{r}
k6.card <- card(k6)
ggplot() +
  geom_histogram(aes(x=k6.card), binwidth = .01) +
  xlab("Number of Neighbors")
```


In contrast, the connectivity graph, clearly demonstrates how each point is connected
to six other points. In our example, this yields a fully connected graph instead of
the collection of sub-graphs for the distance band.

```{r}
plot(k6, coords, lwd=.2, col="blue", cex = .5)
```

## Generalizing the Concept of Contiguity {-}

In GeoDa, the concept of contiguity can be generalized to point layers by converting
the latter to a tessellation, specifically Thiessen polygons. Queen or rook contiguity
weights can then be created for the polygons, in the usual way.

Similarly, the concepts of distance-band weights and k-nearest neighbor weights can be
generalized to polygon layers. The layers are represented by their central points and
the standard distance computations are applied.

We can apply these concepts through R computations, in a similar manner as GeoDa, but
it will take more steps.

### Contiguity-based weights for points {-}


#### Thiessen polygons {-}

An alternative solution to deal with the problem of the uneven distribution of
neighbor cardinality for distance-band weights is to compute a measure of contiguity.
This is accomplished by turning the points into Thiessen polygons. These are also
referred to as Voronoi diagrams or Delaunay triangulations.

In general terms, a Thiessen polygon is a tessellation (a way to divide an area into
regular subareas) that encloses all locations that are closer to the central point
than to any other point. In economic geography, this is a (simplistic) notion of a
market area, in the sense that all consumers in the polygon would patronize the seller
located at the central point. The polygons are constructed by combining lines
perpendicular at the midpoint of a line that connects a point to its nearest
neighbors. From this, the most compact polygon is created.



#### Creating Thiessen polygons from a point layer {-}


To create the Thiessen polygons, we will use the **deldir** package, then will
convert them to a data form that we can use for the contiguity weights. We won't go into
too much detail about the **deldir** package, other than how to create the polygons
and convert them to **sp**, then **sf**. For more information on the **deldir**, visit
[deldir documentation](https://cran.r-project.org/web/packages/deldir/deldir.pdf).

To create our Thiessen polygons from the `deldir` function, we just need a vector 
with x coordinates and a vector with y coordinates. From there we can use the `plot`
command to visualize these because here is a built in method for **deldir** Thiessen polygons. To get a good visual of our polygons, we need a few extra parameters.
These are: `wlines`, `wpoints`,  and `lty`. `wlines = "tess"` gets us the
basic polygons. `wpoints = "none"` keeps points off the map. `lty = 1` specifies
a solid line type.


```{r}
vtess <- deldir(clev.points$x, clev.points$y)
plot(vtess, wlines="tess", wpoints="none",
      lty=1)
```


The function below converts class **deldir** to **sp**. We will not go in depth on the
structure and set up of this function because it goes to heavy into **sp**, and these
notebooks are primarily focused on **sf**. 
```{r}
voronoipolygons = function(thiess) {
w = tile.list(thiess)
polys = vector(mode='list', length=length(w))
for (i in seq(along=polys)) {
    pcrds = cbind(w[[i]]$x, w[[i]]$y)
    pcrds = rbind(pcrds, pcrds[1,])
    polys[[i]] = Polygons(list(Polygon(pcrds)), ID=as.character(i))
}
SP = SpatialPolygons(polys)
 voronoi = SpatialPolygonsDataFrame(SP, data=data.frame(dummy = seq(length(SP)), row.names=sapply(slot(SP, 'polygons'), 
    function(x) slot(x, 'ID'))))
}
```


The result of our function is a `SpatialPolygonsDataFrame`, which is a class from the
**sp** package. We will convert again to get to the **sf** class, since **sf** is more 
efficient and accurate. We check our function by plotting the result and get something
similar to what we plotted earlier, just now it is a different data structure.

```{r}
v <- voronoipolygons(vtess)
plot(v)
```

Using the **sf** function `st_as_sf`, we can convert to **sf** from our `SpatialPolygonsDataFrame`.
Again, we use the plot command to check our result.

```{r}
vtess.sf <- st_as_sf(v)
plot(vtess.sf$geometry)
```

### Contiguity weights for Thiessen polygons {-}


To make our queen contiguity weights, we make a function using `st_relate` and 
specifying a DE9-IM pattern. For queen contiguity, our pattern is `"F***T****"`. We
don't really need to go into why the pattern is what it is for our purposes, but
[DE9-IM](https://en.wikipedia.org/wiki/DE-9IM) and [st_relate documentation](https://www.rdocumentation.org/packages/sf/versions/0.6-3/topics/st_relate) can help explain this to a degree.

```{r}
st_queen <- function(a, b = a) st_relate(a, b, pattern = "F***T****")
```


Here we use our function to get a neighbor list of class **sgbp**, which stands for
sparse geometry binary predicate. In order to use the **spdep** library, we will have to
convert to class **nb**, as done in the contiguity weights notebook. 

```{r}
queen.sgbp <- st_queen(vtess.sf)
```


Now we will we start our conversion from class **sgbp** to **nb**. To do this,
we need to change the class explicitly and take the precaution to represent 
observations with no neighbors with the integer 0. Our data set doesn't have
any observations without neighbors, but in ones with these, it will mess everything
up, if not dealt with. 

We start with the `function` operator. The input for our function will be an object
of class **sgbp**, denoted by **x**. We store the attributes in **attrs**, as we will
need to reapply them later. Now we deal we observation with no neighbors. We will
use `lapply`, which applies a function to each element of a list or vector. As for the
input function, we make one that checks the length of an element of our list for
0(meaning no neighbors) and returns 0 if the element is empty and the element
otherwise. This can be a bit confusing, for more information on `lapply`, check out 
[lapply documentation](https://www.rdocumentation.org/packages/base/versions/3.5.1/topics/lapply)

From here we will have dealt with observations with no neighbors, but will need to 
reapply our attributes to the resulting structure from the `lapply` function. This 
is done by calling the `attributes` function and assigning the stored attributes.
The final step is to explicitly change the class to **nb** by using the `class` function
and assigning `"nb"`. We then return our object **x**.

```{r}
as.nb.sgbp <- function(x, ...) {
  attrs <- attributes(x)
  x <- lapply(x, function(i) { if(length(i) == 0L) 0L else i } )
  attributes(x) <- attrs
  class(x) <- "nb"
  x
}
```

With our conversion function, we can covert to class **nb**.

```{r}
queen.nb <- as.nb.sgbp(queen.sgbp)
```


We can take a look at the distribution of the number of neighbors in our queen contiguity
for the Thiessen polygons with a connectivity histogram. The steps are the same as earlier
for all the other connectivity histograms


```{r}
queen.nb.card <- card(queen.nb)
ggplot() +
  geom_histogram(aes(x=queen.nb.card)) +
  xlab("Number of Neighbors")
```

```{r}
summary(queen.nb)
```



The histogram and summary statistics represents a much more symmetric and compact
distribution of the neighbor cardinalities, very similar to the typical shape of the
histogram for first order contiguity between polygons. The median number of neighbors is 6
and the average 5.6, with a limited spread around these values. In many instances where the
point distribution is highly uneven, this approach provides a useful compromise between the
distance-band and the k-nearest neighbors.

This will be further illustrated through a connectivity graph with a far more balanced structure.


```{r}
plot(queen.nb,coords, lwd=.2, col="blue", cex = .5)
```



## Distance-based weights for polygons {-}



As we can do contiguity-based weights for point data, we use distance-based weights 
for polygons. To illustrate this, we will use U.S homicide data from a previous
notebook(the contiguity-based weights notebook).

To implement this we start by computing shape center coordinates for the
polygons. Once we get the coordinates, we will follow the procedure for making
distance-band weights. There will be one key difference, as we will be working
with nonprojected data. We will have to use great circle distance instead of
euclidean. The formulas for these were went over earlier, in the notebook, but 
working with this difference requires only a few changes in the steps to make
distance-band neighbors.


### Getting the data {-}



To get the data, we will use the geodaData package again. 
Additionally, the data for this notebook can be found at [US Homicides](https://geodacenter.github.io/data-and-lab/ncovr/).
If the data is downloaded directly, then you must put it in your working directory
and load it with `st_read`.

```{r}
us.bound <- geodaData::ncovr
```



### Computing the centroids {-}

Getting our coordinate data is a little more complicate than just using `st_centroid`
on our shapefile. We need to do this to get coordinates in a supported format for
the neighbor functions in the *spdep** package. We are going to get vectors for
latitude and longitude, then will bind the columns for later use.

To get our centroid coordinates, we will map the function `st_centroid` over the
geometry column of **us.bound**. The mapping function applies a given function to each 
element of a vector and returns a vector of the same length. Our input vector will be 
the geometry column of **us.bound**. Our function will be `st_centroid`. We will use the 
`map_dbl` variation of `map` from the **purrr** package. For more documentation, check out [map documentation](https://www.rdocumentation.org/packages/purrr/versions/0.2.5/topics/map)

```{r}
latitude <- map_dbl(us.bound$geometry, ~st_centroid(.x)[[1]])
```

We do the same for latitude, using `[[2]]` to get the latitude values.

```{r}
longitude <- map_dbl(us.bound$geometry, ~st_centroid(.x)[[2]])
```

Now we bind the coordinates together with `cbind`, so we can enter them as parameters
in the neighbors functions.

```{r}
center.coords <- cbind(latitude,longitude)
```




### Computing the critical threshold {-}

To get the critical threshold, we follow the same steps as earlier, with one difference.
This time we will set `longlat = TRUE` in three of the functions. This is important because our data is not
projected and to get accurate results we must work with great circle distance.

Again we begin with finding the first nearest neighbors and convert from class **knn** to class **nb**, and 
set `longlat = TRUE`.

```{r}
k.poly <- knn2nb(knearneigh(center.coords, longlat = TRUE))
```


After we have computed the first nearest neighbors, we find the maximum distance in 
this neighbors list, so our distance-band weights will not have isolates. For our distance
function: `nbdists` we also need to set `longlat = TRUE`. This is needed here, as we are calculating
the distances between each set of points.

```{r}
critical.threshold.poly <- max(unlist(nbdists(k.poly,center.coords, longlat = TRUE)))
```




### Computing distance-band neighbors {-}

We now have all the components necessary to do the distance-band weight for our polygon data.
Again we use `dnearneigh` to get our distance-band weights. We input the centroid coordinates,
a distance lower bound(0), a distance upper bound(the critical threshold), and set `longlat = TRUE`.


```{r}
nb.dist.band.poly <- dnearneigh(center.coords, 0, critical.threshold.poly, longlat = TRUE)
```


Once we have calculated the distance-band weights, we can assess the distribution of the number of 
neighbors through a connectivity histogram, following the same procedure used through the notebook.

```{r}
poly.nb.card <- card(nb.dist.band.poly)
ggplot() +
  geom_histogram(aes(x=poly.nb.card)) +
  xlab("Number of Neighbors")
```



