---
title: "Local Spatial Autocorrelation 2"
author: Luc Anselin and Grant Morrison^[University of Chicago, Center for Spatial
  Data Science -- anselin@uchicago.edu,morrisonge@uchicago.edu]
date: "06/30/2019"
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_document:
    css: tutor.css
    fig_caption: yes
    self_contained: no
    toc: yes
    toc_depth: 4
subtitle: R Notes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

## Introduction {-}

This notebook cover the functionality of the [Local Spatial Autocorrelation](https://geodacenter.github.io/workbook/6a_local_auto/lab6a.html) section of the GeoDa workbook. We refer to that document for details on the methodology, references, etc. The goal of these notes is to approximate as closely as possible the operations carried out using GeoDa by means of a range of R packages.

The notes are written with R beginners in mind, more seasoned R users can probably skip most of the comments
on data structures and other R particulars. Also, as always in R, there are typically several ways to achieve a specific objective, so what is shown here is just one way that works, but there often are others (that may even be more elegant, work faster, or scale better).

For this notebook, we use Cleveland house price data. Our goal in this lab is show how to assign spatial weights based on different distance functions.


```{r}

```
### Objectives

After completing the notebook, you should know how to carry out the following tasks:

- Identify clusters with the Local Moran cluster map and significance map

- Identify clusters with the Local Geary cluster map and significance map

- Identify clusters with the Getis-Ord Gi and Gi* statistics

- Identify clusters with the Local Join Count statistic

- Interpret the spatial footprint of spatial clusters

- Assess potential interaction effects by means of conditional cluster maps

- Assess the significance by means of a randomization approach

- Assess the sensitivity of different significance cut-off values

- Interpret significance by means of Bonferroni bounds and the False Discovery Rate (FDR)

#### R Packages used

- **sf**: To read in the shapefile and make queen contiguity weights

- **spdep**: To create spatial weights structure from neighbors structure

- **robustHD**: To compute standarized scores for variables and lag variables 

- **tmap**: To construct significance and cluster maps with custom functions

- **tidyverse**: To manipulate the data

- **RColorBrewer**: To create custom color palattes that mirror the GeoDa cluster and significance maps

#### R Commands used

Below follows a list of the commands used in this notebook. For further details
and a comprehensive list of options, please consult the 
[R documentation](https://www.rdocumentation.org).

- **Base R**: `install.packages`, `library`, `setwd`, `summary`, `attributes`, `lapply`, `class`, `length`, `rev`, `cut`, `mean`, `sample`, `as.data.frame`, `matrix`, `unique`, `as.character`, `which`, `order`, `data.frame`, `ifelse`, `sum`, `rep`, `set.seed`

- **sf**: `st_read`, `st_relate`

- **spdep**: `nb2listw`, `lag.listw`

- **robustHD**: `standardized`

- **tmap**: `tm_shape`, `tm_borders`, `tm_fill`, `tm_layout`, `tm_facets`

- **tidyverse**: `filter`, `mutate`

- **RColorBrewer: `brewer.pal`

## Preliminaries

Before starting, make sure to have the latest version of R and of packages that are compiled for the matching version of R (this document was created using R 3.5.1 of 2018-07-02). Also, optionally, set a working directory, even though we will not
actually be saving any files.^[Use `setwd(directorypath)` to specify the working directory.]

### Load packages

First, we load all the required packages using the `library` command. If you don't have some of these in your system, make sure to install them first as well as
their dependencies.^[Use 
`install.packages(packagename)`.] You will get an error message if something is missing. If needed, just install the missing piece and everything will work after that.
```{r}
library(sf)
library(spdep)
library(tmap)
library(tidyverse)
library(RColorBrewer)
library(robustHD)
library(geodaData)
```

### geodaData

All of the data for the R notebooks is available in the **geodaData**
package. We loaded the library earlier, now to access the individual
data sets, we use the double colon notation. This works similar to
to accessing a variable with `$`, in that a drop down menu will 
appear with a list of the datasets included in the package. For this
notebook, we use `guerry`.


```{r}
us.homocides <- geodaData::ncovr
```




### Making the weights



To start we create a function for queen contiguity, which is just `st_relate` with
the specified pattern for queen contiguity which is `F***T****`
```{r}
st_queen <- function(a, b = a) st_relate(a, b, pattern = "F***T****")
```


We apply the queen contiguity function to the guerry polygons and see that the class
of the output is **sgbp**. This structure is close to the **nb** structure, but has
a few difference that we will need to correct to use the rest of **spdep** functionality.
```{r}
queen.sgbp <- st_queen(us.homocides)
class(queen.sgbp)
```


This function converts type **sgbp** to **nb**. It is covered in more depth in the 
Contiguity Based Weight notebook. In short, it explicitly changes the name of the 
class and deals with the observations that have no neighbors.
```{r}
as.nb.sgbp <- function(x, ...) {
  attrs <- attributes(x)
  x <- lapply(x, function(i) { if(length(i) == 0L) 0L else i } )
  attributes(x) <- attrs
  class(x) <- "nb"
  x
}
```





```{r}
queen.nb <- as.nb.sgbp(queen.sgbp)
```

To go from neighbors object to weights object, we use `nb2listw`, with default parameters, we will 
get row standardized weights.
```{r}
queen.weights <- nb2listw(queen.nb, style = "B", zero.policy = TRUE)
```


## Extensions of Local Moran's I

### Bivariate Local Moran's I

#### Concept

The treatment of the bivariate local Moran’s I closely follows that of its global counterpart (see also
Anselin, Syabri, and Smirnov 2002). In essence, it captures the relationship between the value for one
variable at location i, $x_i$, and the average of the neighboring values for another variable, i.e,
it spatial lag $\Sigma_jw_{ij}y_j$. Apart from a constant scaling factor (that can be ignored), the
statistic is the product of $x_i$ with the spatial lag of $y_i$ (i.e, $\Sigma_jw_{ij}y_j$), with 
both variables standardized, such that their means are zero and variences equal 1.

$$I_{B,i} = cx_i\Sigma_jw_{ij}y_j$$,

where $w_{ij}$ are elements of the spatial weights matrix.

As for its global counterpart, this statistic needs to be interpreted with caution, since it ignores
in-situ correlation between the two variables (see the discussion of global bivariate spatial
autocorrelation for details).


A special case of the bivariate local Moran statistic is comparing the same variable at two points in time. The most meaningful application is where one variable is for time period t, say $z_t$ and
the other variable is for the neighbors in the previous time period, say $z_{t-1}$. This is illustrated
in the example below. 


#### Implementation


##### Computing the observed statistic


To start we will make a function that calculates the observed bivariate local moran
statistic. There are many ways to approach this, but it will be most efficient to
to work with a full spatial weights matrix rather than the reduced structure 
utilized by **spdep**. This is because it is much faster when applying the conditional
randomization approach in assessing significance.

Before proceeding we will turn the weights structure into a **symmetricMatrix**. We
can do this with `as`, our queen weights and "symmetricMatrix" as arguments. Next
we use `as.matrix` with **W/rowSums(W)** as a precaution to make sure our weights 
are row-standardized. Lastly we assign 0 to all NA's.
```{r}
W  <- as(queen.weights, "symmetricMatrix")
W  <- as.matrix(W/rowSums(W))
W[which(is.na(W))] <- 0
```


In the function, we must first standardize the x and y variables. We use `standardize`
to accomplish this. To get the statistic, we follow the formual from above, but
use matrix multiplication. Multiplying the full spatial weight matrix by the standardized
y variable still gets us the standardized lag variable for y. To get matrix multiplication,
we use `%*%`.

```{r}
bivariate_moran <- function(x, y = NULL, W){
        if(is.null(y)) y = x
        xs <- standardize(x)
        ys <- standardize(y)
        local  <- (xs*W%*%ys)
        local
}
```


Here we use the function made above on our x and y variables.
```{r}
lmoran <- bivariate_moran(us.homocides$HR90, us.homocides$HR80,W)
```



##### Conditional permutation


In order to assess significance of the local statistics, we will need to do some conditional
permutations. With these, we can obtain a pseudo p-value or determine a cutoff for 
significance.

Before we proceed with the conditional permutation process, we set the seed to `123456789`.
This is the same randomization seed as GeoDa. Randomization seeds are used for replicability
in research.
```{r}
set.seed(123456789)
```

We will make a function to do the conditional permutations for the bivariate moran. We 
will use a full size spatial weights matrix to take adavtantage of the speed of 
vectorized operation in R. 

To begin, we need to number of locations(**n**). We use `nrow` on the weights matrix, but 
could also use `ncol` since it is a full size spatial weights matrix, which is nxn.
Next we create a vector of id numbers. We will need a place to store the **y** result of
each sample and statistic. **local.sims** is for the statistics and **y.sample** is for
the y sample draws. Now we need a `for` loop in the sample draws because we must draw
samples from all of the locations except location i in the condional permutations. To
accomplish this, we have to loop through 1:n and take a sample of length **permutations**
from **id[-i]**. The minus notation excludes index i from the vector, so we are taking a
sample of all locations except i. We take a sample of id numbers, so we can index the
**y** vector to get the sample values. at the end of the loop we have a full matrix
of **y** sample values for each location and permutation. We then use `apply` to standardize
the y values in the matrix.


```{r}
bivariate_moran_sim <- function(x, y = NULL, W, permutations = 999){
        if(is.null(y)) y = x

        n   <- nrow(W)
        id  <- 1:n

        #place to store results
        local.sims  <- matrix(NA, nrow = n, ncol=permutations)
        y.sample = matrix(NA, nrow = n, ncol = permutations)
        
        # filling each column of the sample matrix
        for(i in 1:n){
          sample.indices <- sample(id[-i], permutations, replace = TRUE)
          y.sample[i,] <- y[sample.indices]
        }
        
        #standardizing the y sample values
        y.sample <- (y.sample - apply(y.sample, 1, mean))/apply(y.sample, 1, sd)
        
        #standardizing x
        xs <- standardize(x)
        W[which(is.na(W))] <- 0
        
        #calculating the local statistics
        local.sims  <- (xs*W%*%y.sample)
        local.sims
}
```

We use the function from above on our x and y variables.
```{r}
local.sims <- bivariate_moran_sim(us.homocides$HR90,us.homocides$HR80, W)
```



To assess significance at the .05 level for a two-sided test as in GeoDa, we will us an
alpha level of .1. `probs` will be .05 and .95, which corresponds with a two sided test
at the .05 level. 
```{r}
alpha <- .1  # for a 90% confidence interval
probs <- c(alpha/2, 1-alpha/2)
intervals <- t( apply(local.sims, 1, function(x) quantile(x, probs=probs)))
sig        <- (lmoran < intervals[,1] )  | ( lmoran > intervals[,2] )
```

##### Classification


To build the classifications, we will need the standardized scores for both the x and y variables. We
build the classifications in the same manner as the univariate Local Moran. The quadrant of the
moran scatterplot gives us the cluster classification. Positive values give the designation "High",
and negative values "Low"
```{r}
xs <- standardize(us.homocides$HR90)
ys <- standardize(us.homocides$HR80)
```

We can create the classifications using `interaction, which will assign a boolean based on each 
conditional for each observation. We convert these booleans to strings, so we can use 
`str_replace_all` to get "High" and "Low" in place of TRUE and FALSE. Lastly, we assign 
"Not significant" to all insignificant observations based on the conditional permutation approach
taken above.


```{r}
patterns <- as.character( interaction(xs > 0, W%*%ys > 0) ) 
patterns <- patterns %>% 
        str_replace_all("TRUE","High") %>% 
        str_replace_all("FALSE","Low")
patterns[sig==0] <- "Not significant"
```


Here we add the classification variable, **patterns** to the data frame with `mutate` from 
**tidyverse**.

```{r}
us.homocides <- us.homocides %>% mutate(patterns = patterns)
```



##### Constructing the map




```{r}
pal <- c("#DE2D26","#FCBBA1","#C6DBEF", "#3182BD", "#D3D3D3")
pal
```


```{r}
tm_shape(us.homocides) +
  tm_fill("patterns", palette = pal) +
  tm_borders() +
  tm_layout(main.title = "Bivariate Local Moran's I")
```



##### Putting it all together

We can put together all of the steps above to create a bivariate LISA map function.
The only addition we need is something to match up the colors to the classifications
present in the data. For instance "high-low" or one of the other classifications might 
not always be there for a variable, so to ensure that the mapping function displays
a map with the correct color scheme, we need to check the classifications and construct
the palette based on the categories present. This is done below by making a matrix
matching the colors to the classifications then checking the classification present
in the data with `unique`. With the resulting vector from `unique`, we use the `%in%`
operator on the vector with the classification, which gives us a vector logical
values. From there we use the logical value to select rows in the matrix. With this
we can select the row with the colors as our palette.

```{r}
bivariate_lisa_map <- function(area,x, y = NULL, weights, nperms, sig_cutoff){
  # function to create a bivariate LISA map
  # arguments:
  #    area: sf dataframe
  #    x: vectors of x values
  #    y: a vector of y values
  #    nperms: the number of permuations used to calculate the pvalues
  #    sig_cutoff: the alpha level required for significance
  # returns:
  #    a LISA map in GeoDa style
  if(is.null(y)) y = x
  
  # Converting the weights
  W <- as(weights, "symmetricMatrix")
  W  <- as.matrix(W/rowSums(W))
  W[which(is.na(W))] <- 0
  
  # Computing the observed statistics and reference distributions
  observed <- bivariate_moran(x,y = y,W)
  sims <- bivariate_moran_sim(x,y = y,W, permutations = nperms)
  
  # Assessing significance
  alpha <- sig_cutoff  # for a 90% confidence interval
  probs <- c(alpha/2, 1-alpha/2)
  intervals <- t( apply(sims, 1, function(x) quantile(x, probs=probs)))
  sig        <- ( observed < intervals[,1] )  | ( observed > intervals[,2] )
  
  # Standardizing the x and y vars
  xp <- standardize(x)
  yp <- standardize(y)
  
  #Assigning classifications
  patterns <- as.character( interaction(xp > 0, W%*%yp > 0) ) 
  patterns <- patterns %>% 
        str_replace_all("TRUE","High") %>% 
        str_replace_all("FALSE","Low")
  patterns[sig==0] <- "Not significant"
  
  # Adding the classifications to the data frame
  area <- area %>% mutate(patterns = patterns)
  
  # Constructing the correct palette based on presense of classifications
  classes <- unique(patterns)
  patts <- c("High.High", "High.Low", "Low.High", "Low.Low", "Not significant")
  colors <- c("#DE2D26","#FCBBA1","#C6DBEF", "#3182BD", "#D3D3D3")
  mat <- matrix(c(patts,colors), ncol = 2)
  logi <- patts %in% classes
  pre_col <- matrix(mat[logi], ncol = 2)
  pal <- pre_col[,2]
  
  # Making the map
  tm_shape(us.homocides) +
  tm_fill("patterns", palette = pal) +
  tm_borders() 
}
```







```{r}
significance_map <- function(polys, pvalue_vector, permutations, sig = .05){
  # function to create significance map
  # arguments:
  #    polys: sf dataframe
  #    pvalue_vector: a vector of p-values
  #    permutations: the number of permuations used to calculate the pvalues
  #    min.sig: the alpha level required for significance
  # returns:
  #    a significance map in GeoDa style
  
  target_p <- 1 / (1 + permutations)
  potential_brks <- c(.00001, .0001, .001, .01)
  brks <- potential_brks[which(potential_brks > target_p & potential_brks < sig)]
  brks2 <- c(target_p, brks, sig)
  labels <- c(as.character(brks2), "Not Significant")
  brks3 <- c(-.000001, brks2, 1)
  
  cuts <- cut(pvalue_vector, breaks = brks3,labels = labels)
  polys <- polys %>% mutate(significance = cuts)
  
  
  pal <- rev(brewer.pal(length(labels), "Greens"))
  pal[length(pal)] <- "#D3D3D3"
  
  tm_shape(polys) +
    tm_fill("significance", palette = pal) +
    tm_borders() +
    tm_layout(title = "significance map")
}


get_p_value <- function(mat, observed) {
  nperm <- ncol(mat)
  nlocs <- nrow(mat)
  p_value <- rep(NA,nlocs)
  for(i in 1:nlocs){
    num_greater <- length(which(df[i,] > observed))
    p_value[i] <- (num_greater + 1) / (nperm + 1)
  }
  p_value
}

```




#### Interpretation

As mentioned, the interpretation of the bivariate local Moran cluster map warrants some
caution, since it does not control for the correlation between the two variables at each
location (i.e., the correlation between $x_i$ and $y_i$).  To illustrate this issue, 
consider the space-time case, where one can interpret the results of the high-high and
low-low clusters in Figure 3 as locations where high/low values at time t (in our example,
homicide rates in 1990) were surrounded by high/low values at time t−1 (homicide rates in
1980) more so than would be the case randomly. This could either be attributed to the
surrounding locations in the past affecting the central location at the current time (a form
of a diffusion effect), but could also be due to a strong inertia, in which case there would
be no diffusion. More precisely, if the homicide rates in all locations are highly 
correlated over time, then if the surrounding values in t−1 were correlated with the value
at i in t−1, then they would also tend to be correlated with the value at i in t (through
the correlation between $y_{j,t-1}$ and $y_{j,t}$). Hence, while these findings could be
compatible with diffusion, this is not necessarily the case. The same complication affects 
the interpretation of the bivariate local Moran coefficient between two variables at the 
same point in time.





### Differential Local Moran's I


#### Concept


The differential local Moran statistic is the local counterpart to the differential Moran
scatter plot, discussed in an earlier chapter. Instead of using the observations on a
variable at two different time periods, this statistic is based on the change over time,
i.e., the difference between $y_t$ and $y_{t-1}$. Note that this is the actual difference
and not the absolute difference, so that a positive change will be viewed as high, and a
negative change as low. The differences are used in standardized form, i.e., they are not
the differences between the standardized variable at two points in time, but the
standardized differences between the original values for the variable.


The formal expression for this statistic follows the same logic as before, and consists of
the cross product of the difference between $y_t$ and $y_{t-1}$ at i with the associated
spatial lag:

$$I_{D,i} = c(y_{i,t} - y_{i,t-1})\Sigma_jw_{ij}(y_{j,t} - y_{j,t-1})$$

The scaling constant c can be ignored. Inference is based on conditional permutation. All
the usual caveats hold about multiple comparisons, etc.

#### Implementation

To start we need the difference between the two temporal variables, which we will use
as the variable for a univariate local moran.
```{r}
us.homocides$differ <- us.homocides$HR90 - us.homocides$HR80
```

To make the differential Local Moran's I map, we use the `bivariate_lisa_map` function
that we created earlier in the notebook. This function works for the univariate case,
we just need to enter the x variable as the y as well and it will output the
univariate LISA map. Additionally, we can use the `+` operator and additonal **tmap**
function with this function. Below we use `tm_layout` to give the map a title.
```{r}
bivariate_lisa_map(us.homocides, us.homocides$differ, y =us.homocides$differ,queen.weights,999,.05 ) +
  tm_layout(main.title = "Differential Local Moran's I")
```

#### Interpretation

The first aspect of the results is a much smaller number of significant locations compared
to the univariate and bivariate cluster maps. These are locations where the change in the
variable over time is matched by similar/dissimilar changes in the surrounding locations. It
is important to keep in mind that the focus is on change, and there is no direct connection
to whether this changes is from high or low values.

Two situations can be distinguished, depending on whether the change variable takes on both
positive and negative values, or when all the changes are of the same sign (i.e., all
observations either increase or decrease over time).

When both positive and negative change values are present, the high-high locations will tend
to be locations with a large increase (positive change), surrounded by locations with
similar large increases. The low-low locations will be observations with a large decrease
(negative change), surrounded by locations with similar large decreases. Spatial outliers
will be locations where an increase is surrounded by a decrease and vice versa.

When all changes are of the same sign, the interpretation of high-high and low-low depends
on the sign. Due to the standardization, large positive values will be considered high
(above the mean), whereas large negative values will be labeled low (below the mean). This
should be kept in mind when interpreting the results.


### Local Moran with EB Rate

#### Concept


The last of the extensions of the local Moran’s I pertains to the special case where the
variable of interest is a rate or proportion. As discussed for the Moran scatter plot, the
resulting variance instability can cause problems for the Moran statistic. The EB
standardization suggested by Assunção and Reis (1999) for the global case can be extended to
the local statistic in a straightforward manner. The statistic has the usual form, but is
computed for the standardized rates, z.

$$I_{EB,i} = cz_i\Sigma_jw_{ij}z_j$$

The standardization of the raw rate $r_i$ is the same as before, and is repeated here for
completeness (for a more detailed discussion, see the relevant chapter):


$$z_i = \frac{r_i-\beta}{\sqrt{\alpha+(\beta/P_i)}}$$

with $\beta$ as an estimate of the mean and the denominator as an estimate of the 
standard error.

All inference and interpretation is the same as for the univariate case and is not further
pursued here.



#### Implementation

To get the EB standardized score each location, we will need to know $\beta$ and $\alpha$
are. These are covered in the second global spatial autocorrelation notebook, but since
we have to calculate them explicitly here we will go through the formula for both.

$\beta$ is the total event count over the total population,

$$\beta = \Sigma_iO_i/\Sigma_iP_i$$

$$\alpha = [\Sigma_iP_i(r_i-\beta)^2]/P - \beta/(P/n)$$


To implement this, we will need to use base R vectorized operations to calculate the standardized
scores. To start we calculate each of the variables necessary for computing $\alpha$.

```{r}
beta <- sum(us.homocides$HC90) / sum(us.homocides$PO90)
r_i <- us.homocides$HC90 / us.homocides$PO90
P_i <- us.homocides$PO90
n <- ncol(us.homocides)
P <- sum(us.homocides$PO90)
```

With the variables computed above, we have everything from the formula for $\alpha$.
```{r}
alpha <- sum(P_i * (r_i - beta)^2) / P - beta/(P/n)
```

Now that we have $\alpha$ we can compute the standard error.
```{r}
se <- sqrt(alpha + beta/P_i)
```

Lastly we compute the standardized scores based on the formula from earlier.
```{r}
z_i <- (r_i - beta) / se
```




```{r}
bivariate_lisa_map(us.homocides, z_i, y =z_i,queen.weights,999,.05 ) +
  tm_layout(main.title = "EB Local Moran's I")
```





## Bivariate and Multivariate Local Join Count Statistics


In addition to introducing a univariate local Join Count statistic, Anselin and Li (2019)
also include extensions to a bivariate and multivariate setting. In each instance, the
variables under consideration, say x and z, only take on a value of 0 and 1.

We distinguish between two cases, referred to as co-location and no co-location. In the
first, it is possible for both variables to take the same value at a location. In other
words, at location i, it is possible to have $x_i$ = $z_i$ = 1. For example, this would be
the case when the variables pertain to two different types of events that can occur in all
locations, such as the presence of several non-exclusive characteristics (e.g., a city block
that is both low-density and commercial). In the second instance, whenever $x_i = 1$, then
$z_i = 1$ and vice versa. In other words, the two events cannot happen in the same location.
This would be case where the classification of a location is exclusive, i.e., if a location
is of one type, it cannot be of another type (e.g., a zoning classification).

In GeoDa, the co-location case is treated under Multivariate Local Join Count (which
includes the bivariate setup as a special case), whereas the no co-location case is
implemented under Bivariate Local Join Count.

While the natural application of Join Count statistics is to deal with binary variables,
they can also be extended to a notion of quantile spatial autocorrelation, which is further
elaborated upon below.


#### Concept

In its general form, the expression for a bivariate local join count statistic is

$$BJC_i = x_i(1-z_i)\Sigma_jw_{ij}z_j(1-x_j)$$

with $w_{ij}$ as unstandardized (binary) spatial weights.

The roles of x and z may be reversed, but the statistic is not symmetric, so that the
results will tend to differ whether x or z is the focus. In addition, it is important to
keep in mind that the statistic will only be meaningful when the proportion of the second
variable (for the neighbors) in the population is small.

Since the condition of no co-location ensures that $1-z_i = 0$ whenever $x_i = 1$, and
vice versa, the statistic simplifies to:

$$BJC_i = x_i\Sigma_jw_{ij}z_j$$



#### Implementation





#### Interpretation






### Bivariate and Multivariate - Co-Location





#### Preliminaries


##### Getting the data from the GeoDa website

To get the data for this notebook, you will and to go to [Guerry](https://geodacenter.github.io/data-and-lab/Guerry/) The download format is a
zipfile, so you will need to unzip it by double clicking on the file in your file
finder. From there move the resulting folder titled: nyc into your working directory
to continue. Once that is done, you can use the **sf** function: `st_read()` to read
the shapefile into your R environment. 


```{r}
guerry <- st_read("guerry/guerry.shp")
```




##### Exploratory maps

In this section we will select the top quintile of each variable for the construction of our
binary variables. To get an illustration of these location, we will make basic choropleth maps
with the `quantile` style for the variables **Donatns** and **Infants**


```{r}
tm_shape(guerry) +
  tm_fill("Infants", n = 5, style = "quantile") +
  tm_borders() +
  tm_layout(legend.outside = TRUE)
```



```{r}
tm_shape(guerry) +
  tm_fill("Donatns", n = 5, style = "quantile") +
  tm_borders() +
  tm_layout(legend.outside = TRUE)
```


To get the break for the top quintile of both variables, we use `quantile` with the variables as 
inputs and .8 to get the top quintile break, or 80th percentile.
```{r}
doncatbreak <- quantile(guerry$Donatns, .8)
infantbreak <- quantile(guerry$Infants, .8)
```

Next, we create two vectors to store the binary variables by using `rep` and starting with all
0 values. We use conditional indexing to assign 1 to all values that are greater than the
80th percentile break that we computed above.
```{r}
doncat <- rep(0,85)
infcat <- rep(0,85)
doncat[guerry$Donatns > doncatbreak] <- 1
infcat[guerry$Infants > infantbreak] <- 1
```

With both binary variables, we can create a co-location variable to get a sense of the potential
significant locations under the bivariate local joint count.

```{r}
guerry$coloc <- rep(0,85)
guerry$coloc[doncat == 1 & infcat == 1] <- 1
```

Here we map the co-location variable in the GeoDa style and see the same co-location output.

```{r}
tm_shape(guerry) +
  tm_fill("coloc", palette = c("white", "blue"), style = "cat") +
  tm_borders()
```







##### Making the weights



```{r}
queen.sgbp <- st_queen(guerry)
queen.nb <- as.nb.sgbp(queen.sgbp)
binary.w <- lapply(queen.nb, function(x) 0*x + 1)
queen.weights <- nb2listw(queen.nb, style = "B",glist = binary.w, zero.policy = TRUE)
```



```{r}
W  <- as(queen.weights, "symmetricMatrix")
W  <- as.matrix(W/rowSums(W))
W[which(is.na(W))] <- 0
```


#### Concept

A bivariate co-location cluster requires that $x_i=z_i=1$ coincides with
neighbors for which also $x_j=z_j=1$. For the bivariate case, the corresponding
local Join Count statistic takes the form: 

$$CLC_i = x_iz_i\Sigma_jw_{ij}x_jz_j$$

with $w_{ij}$ as unstandardized (binary) spatial weights.

As before, a conditional permutation approach can be constructed for those locations
with $x_i=z_i=1$. We draw $k_i$ pairs of observations $(x_j,z_j)$ from the set of
N - 1 (this contains P - 1 observations with $x_j = 1$ and Q - 1 observations
with $z_j=1$). In a one sided test, we again count the number of times the statistic
equals or exceeds the observed join count value at i.

The extension to more than two variables is mathematically straightforward. We
consider m variables at each location i, i.e., $x_{hi}$ for h = 1 ,…, m, with 
$\Pi_{h=1}^mx_{hi} = 1$, which enforces the co-location requirement.

The corresponding statistic is then:

$$CLC_i = \Pi_{h=1}^mx_{hi}\Sigma_jw_{ij}\Pi_{h=1}^mx_{hj}$$

The implementation of a conditional permutation strategy follows as a direct
generalization of the bivariate co-location cluster. However, for a large number of
variables, such co-locations become less and less likely, and a different conceptual
framework may be more appropriate.

#### Quantile Local Spatial Autocorrelation


The local Join Count statistics are designed to deal with binary variables. However, as we
illustrate with our example, they can also be applied to assess the spatial autocorrelation
between quantiles of a continuous variable, something that can be referred to as quantile
local spatial autocorrelation. As mentioned, the continuous (linear) association between
two variables that is measured by the bivariate local Moran suffers from the problem of
in-situ correlation. The quantile local spatial autocorrelation sidesteps this problem by
converting the continuous variable to a binary variable that takes the value of 1 for a
specific quantile. For example, as in our illustration, the spatial assocation between two
continuous variables can be simplified by focusing on the local autocorrelation between
binary variables for the upper quintile. Since the bivariate (or multivariate) local Join
Count statistic enforces co-location, the problem of in-situ correlation is controlled for.
This provides an alternative to other bivariate and multivariate local spatial
autocorrelation statistics, albeit with a loss of information (going from the continuous
variable to a binary category).



#### Implementation


To start we make a function that computes the bivariate join count statistic. This is pretty
simple as all we need is to multiply the two vectors, then compute the statistic in the same way
that we would for the univariate local moran. It is importnat to note that you must have the correct 
set up for this function to yield accurate results. The input weights matrix must be binary and
the x and z variables need to also be binary.
```{r}
bivariate_jc <- function(x,z,W){
  xz <- x*z
  jc <- xz * (W %*% xz)
  jc
}
```




```{r}
x <- doncat
z <- infcat
observed <- bivariate_jc(x,z,W)
```

Next we create a function for the conditional permutations. This follows the same process as the 
one we made for the bivariate local moran simulations. First we get the number of observations
and a vector of id numbers. Next we create data structures to store the sample draws. Then we 
loop through and sample from all the id numbers except at location i as GeoDa does. Lastly
we compute the **xz.sample** matrix, which we will use in the final calculation of the reference
statistics. **xz.sample** is only used for the neighbor portion of the computation, as the value at 
location i of $x_iz_i$ is held constant throughout the permutations.

```{r}
bivariate_join_count_sims <- function(x, z, W, permutations = 999){

        n   <- nrow(W)
        id  <- 1:n

        #place to store results
        local.sims  <- matrix(NA, nrow = n, ncol=permutations)
        x.sample = matrix(NA, nrow = n, ncol = permutations)
        z.sample = matrix(NA, nrow = n, ncol = permutations)
        
        # filling each column of the sample matrices
        for(i in 1:n){
          sample.indices <- sample(id[-i], permutations, replace = TRUE)
          x.sample[i,] <- x[sample.indices]
          z.sample[i,] <- z[sample.indices]
        }
      
        #standardizing x
        xz.sample <- x.sample * z.sample
        xz <- x * z
        
        #calculating the local statistics
        local.sims  <- (xz*W%*%xz.sample)
        local.sims
}
```


```{r}
sims <- bivariate_join_count_sims(x,z, W)
```




With the simulations and the observed local statistics, we will need a function to compute the 
p-values. This is used in the previous notebook on Local Spatial Autocrrelation. For most
of the local spatial statistics, we use a two sided test, but in this case we will be working 
with a one-sided test as in GeoDa.

```{r}
get_p_value <- function(mat, observed,type) {
  nperm <- ncol(mat)
  nlocs <- nrow(mat)
  p_value <- rep(NA,nlocs)
  for(i in 1:nlocs){
    num_greater <- length(which(mat[i,] >= observed[i]))
    p_value[i] <- (num_greater + 1) / (nperm + 1)
  }
  if (type == "two-sided"){
    p_value <- ifelse(p_value > .5, 1-p_value, p_value)
    p_value
  } else {
    p_value
  }
}
```




```{r}
p_values <- get_p_value(sims, observed)
p_values
```


```{r}
significance_map(guerry, p_values,999)
```




### Multivariate Local Geary

#### Preliminaries


#### Concept

The multivariate local Geary statistic is formally the sum of individual local Geary
statistics for each of the variables under consideration. For example, with m variables,
indexed by h, the corresponding expression is:

$$c_i = \Sigma_{h=1}^m\Sigma_jw_{ij}(x_{hi}-x_{hj})^2$$

This measure corresponds to a weighted average of the squared distances in multidimensional
attribute space between the values observed at a given geographic location i and those at its
geographic neighbors j.

Inference can again be based on a conditional permutation approach. This consists of holding
the tuple of values observed at i fixed, and computing the statistic for a number of
permutations of the remaining tuples over the other locations. This results in an empirical
reference distribution that represents a computational approach at obtaining the distribution
of the statistic under the null. The resulting pseudo p-value corresponds to the fraction of
statistics in the empirical reference distribution that are equal to or more extreme than the
observed statistic.

uch an approach suffers from the same problem of multiple comparisons mentioned for all the other local statistics. In addition, there is a further complication. When comparing the
results for m univariate Local Geary statistics, these multiple comparisons need to be
accounted for as well. For example, for each univariate test, the target p-value of
$\alpha$ would typically be adjusted to $\alpha/m$  (with m variables, each with a univariate
test), as a Bonferroni bound. Since the multivariate statistic is in essence a sum of the
statistics for the univariate cases, this would suggest a similar approach by dividing the
target p-value by the number of variables (m). Alternatively, and preferable, a FDR strategy
can be pursued. The extent to which this actually compensates for the two dimensions of
multiple comparison (multiple variables and multiple observations) remains to be further
investigated.

#### Implementation


##### Preliminaries




```{r}
queen.weights <- nb2listw(queen.nb, style = "B", zero.policy = TRUE)
W  <- as(queen.weights, "symmetricMatrix")
W  <- as.matrix(W/rowSums(W))
W[which(is.na(W))] <- 0
```



```{r}
x <- guerry$Donatns
y <- guerry$Infants
x2 <-  guerry$Donatns ^ 2
y2 <- guerry$Infants ^ 2
lg_x <- x2 - 2*x*(W%*%x) + W%*%x2
lg_y <- y2 - 2*y*(W%*%y) + W%*%y2
observed_lg <- lg_x + lg_y
```



```{r}
bivariate_local_geary <- function(x,y,W){
  x2 <- x^2
  y2 <- y^2
  lg_x <- x2 - 2*x*(W%*%x) + W%*%x2
  lg_y <- y2 - 2*y*(W%*%y) + W%*%y2
  lg <- lg_x + lg_y
  lg
} 
```









```{r}
bivariate_local_geary_sims <- function(x,y,W,permutations = 999){
  n <- nrow(W)
  id  <- 1:n
  
  local.sims  <- matrix(NA, nrow = n, ncol=permutations)
  x.sample = matrix(NA, nrow = n, ncol = permutations)
  y.sample = matrix(NA, nrow = n, ncol = permutations)
  
  for(i in 1:n){
    sample.indices <- sample(id[-i], permutations, replace = TRUE)
    x.sample[i,] <- x[sample.indices]
    y.sample[i,] <- y[sample.indices]
  }
  x2 <- x^2
  y2 <- y^2
  x2.sample <- x.sample ^ 2
  y2.sample <- y.sample ^ 2
  
  local.sims <- (x2 - 2*x*W%*%x.sample + W%*%x2.sample) + (y2 - 2*y*W%*%y.sample + W%*%y2.sample)
  local.sims
}
```





```{r}
mlg.sims <- bivariate_local_geary_sims(x,y,W)
p.values <- get_p_value(mlg.sims,observed_lg, type = "two-sided")
```






```{r}
significance_map(guerry,p.values,999)
```





```{r}
mean_lg <- rep(NA, 85)
for (i in 1:85){
  mean_lg[i] <- mean(mlg.sims[i,]) 
}
```





```{r}
patterns <- as.character(interaction(mean_lg > observed_lg))
patterns <- patterns %>% 
        str_replace_all("TRUE","Positive") %>% 
        str_replace_all("FALSE","Negative")
patterns[p.values > .05] <- "Not Significant"
```


```{r}
guerry <- guerry %>% mutate(lg_classifications = patterns)
```


```{r}
tm_shape(guerry) +
  tm_fill("lg_classifications", palette = c("#C6DBEF","#D3D3D3","#3182BD")) +
  tm_borders()
```





```{r}
bivariate_local_geary_map <- function(area, x, y, weights, nperms,sig_cutoff){
  if(is.null(y)) y = x
  
  # Converting the weights
  W <- as(weights, "symmetricMatrix")
  W  <- as.matrix(W/rowSums(W))
  W[which(is.na(W))] <- 0
  
  # Computing the observed statistics and reference distributions
  observed <- bivariate_local_geary(x,y,W)
  sims <- bivariate_local_geary_sims(x,y, W, nperms)
  
  # Assessing significance
  alpha <- sig_cutoff  # for a 90% confidence interval
  probs <- c(alpha/2, 1-alpha/2)
  intervals <- t( apply(sims, 1, function(x) quantile(x, probs=probs)))
  sig <- ( observed < intervals[,1] )  | ( observed > intervals[,2] )
  
  mean_lg <- rep(NA, 85)
  for (i in 1:85){
    mean_lg[i] <- mean(sims[i,]) 
  }
  
  #Assigning classifications
  patterns <- as.character(interaction(mean_lg > observed))
  patterns <- patterns %>% 
        str_replace_all("TRUE","Positive") %>% 
        str_replace_all("FALSE","Negative")
  patterns[sig == 0] <- "Not Significant"
  
  # Adding the classifications to the data frame
  area <- area %>% mutate(lg_classifications = patterns)
  
  
  # Constructing the correct palette based on presense of classifications
  classes <- unique(patterns)
  patts <- c("Positive", "Negative", "Not Significant")
  colors <- c("#C6DBEF","#3182BD","#D3D3D3")
  mat <- matrix(c(patts,colors), ncol = 2)
  logi <- patts %in% classes
  pre_col <- matrix(mat[logi], ncol = 2)
  pal <- pre_col[,2]
  
  # Making the map
  tm_shape(area) +
  tm_fill("patterns", palette = pal) +
  tm_borders() 
}
```


```{r}
tm_shape(guerry) +
  tm_fill("lg_classifications", palette = c("#C6DBEF","#D3D3D3","#3182BD")) +
  tm_borders()
```


```{r}
bivariate_local_geary_map(guerry,x,y,queen.weights,999,.05)
```


#### Interpretation

The results of the multivariate local Geary significance or cluster map are not always easy to
interpret. Mainly, this is because we tend to simply superimpose the results of the univariate tests,
whereas the multivariate statistic involves different tradeoffs.

























